---
title: "💻 Week 07 - Lab Roadmap (90 min)"
date: 6 November 2022
date-meta: 6 November 2022
from: markdown+emoji
author: Dr. Stuart Bramwell/Dr. Jon Cardoso-Silva
---

📥 Download the RMarkdown version of this roadmap [from Moodle](https://moodle.lse.ac.uk/mod/folder/view.php?id=1134401&forceview=1).

-------------------

This lab session draws on [🗓️ Week 05 lecture](/weeks/week05/lecture.qmd) content and on feedback given by the course representatives about the main struggles you are facing with R (tidyverse).

If you are already very familiar with tidyverse, you can skip to Step 2.

**R packages you will need:**

```{r, message=FALSE, warning=FALSE}
library('e1071')
library('ggsci')
library('ISLR2')
library('palmerpenguins')
library('parttree')
library('rpart')
library('rpart.plot')
library('tidyverse')
```

You might have already installed some of these packages. Some were used in [🗓️ Week 05 slides](/weeks/week05/lecture.qmd), others in previous labs. If you see an error like "package not found", then install the package using `install.packages("<package name>")`.

:::{.callout-important collapse="true"}
## Installation instructions for `parttree`

The package parttree cannot be installed by install.packages. Instead, we have to follow the instructions set by the [developers of the package](https://github.com/grantmcdermott/parttree):

```r
install.packages("remotes")
remotes::install_github("grantmcdermott/parttree", force = TRUE)
```
:::


# Step 1: A `tidyverse` recap (40 mins)

_(the lab we wish we had run in the first weeks of this course!)_

## Step 1.1: The tibble (5 min)

So far, we have been working with data frames. We will introduce you to a different kind of data frame called a tibble. Tibbles are like data frames, only they have a more descriptive print function, and you can perform more advanced tasks like looping over lists (without needing to specify a for loop). 

Let's start by converting our now familiar `Boston` data to a tibble using `as_tibble`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
boston <- as_tibble(Boston)
boston
```

Instead of printing a lot of rows, we only get to see the first ten rows. 

We can also see the dimensions of the data `506 x 13` and the class of each variable. 

So with one command, we can get a lot more **useful** information on our data without the need for multiple commands.

## Step 1.2: Basic `dplyr` verbs (10 minutes)

In previous labs, we have been using base R to subset data based on the rows and columns, to create new variables, and to create summary statistics.  However, there is a more verbal way of performing these tasks using the `tidyverse`. We will introduce you to several key verbs, namely `filter`, `select`, `mutate` and `summarise`.

**`Auto` cylinders**

Suppose we wanted to select only the rows of the `Auto` data for cars with 4 cylinders. We can achieve this using the following base R command:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
Auto[Auto$cylinders == 4, ]
```

Now here is the `tidyverse` solution using `filter`.

```{r, eval=TRUE, echo=TRUE, output=FALSE}
filter(Auto, cylinders == 4)
```

**`Boston` columns**

Next, let's only include `lstat` and `medv` from `Boston`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
Boston[, c('medv','lstat')]
```

Now here is the `tidyverse` solution using `select`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
select(Boston, medv, lstat)
```

**`Carseats` median sales**

Now that we can subset variables, let's create some new ones. Let's create a dummy variable `SalesGTMedian` for each car seat in `Carseats`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
Carseats$SalesGTMedian <- if_else(Carseats$Sales > median(Carseats$Sales), TRUE, FALSE)
```

Now here is the `tidyverse` solution using `mutate`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
mutate(Carseats, SalesGTMedian = if_else(Sales > median(Sales), TRUE, FALSE))
```

**Missing data in the `Hitters` dataset**

Finally, suppose we wanted to find the average `Salary` in `Hitters` (a basketball data set). We specify `na.rm = TRUE` to get R to ignore all the missing values.

```{r, eval=TRUE, echo=TRUE, output=FALSE}
mean(Hitters$Salary, na.rm = TRUE)
```

Now here is the `tidyverse` solution using `summarise`.

```{r, eval=TRUE, echo=TRUE, output=FALSE}
summarise(Hitters, mean_salary = mean(Salary, na.rm = TRUE))
```

**Categorical variables in `Default` dataset**

Some of our variables will be categories, so let's find out the distribution of defaults in `Default`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
table(Default$default)
```

Now here is the `tidyverse` solution using `count`:

```{r, eval=TRUE, echo=TRUE, output=FALSE}
count(Default, default)
```

💡 **REFLECTION TIME**

Let's pause for a minute to see the advantages of these commands:

1. the commands themselves give a better indication of what it is we are trying to do. This is highly advantageous when it comes to communicating our code with others. 

2. when working with variables in data frames, we do not need to use `$`. Instead, we can just reference the variable on its own, provided we pass the command the data frame. 

3. **every time we use these <mark>verbs</mark>, a new data frame is created** - meaning we can use the output to create ggplots!

## Step 1.3 The pipe (10 minutes)

You may have seen `%>%` in some of our code. This is known as the <mark>pipe</mark> operator, and it enables us to chain together multiple verbs into one fluid sequence of steps. To get this quickly you can use `ctrl+shift+m` (for Windows users) or `command+shift+m` (for Mac users).

Suppose we want to find out what proportion of American cars had `mpg` above the global average. We can find this out by using the following sequence of commands:

```{r}
Auto %>% 
  as_tibble() %>% 
  select(mpg, origin) %>% 
  mutate(mpg_gt_gavg = if_else(mpg > mean(mpg), TRUE, FALSE)) %>% 
  filter(origin == 1) %>% 
  summarise(prop_mpg_gt_gavg = mean(mpg_gt_gavg))
```

Let's walk through what we just did:

1. We converted `Auto` to a tibble.
2. We selected only the variables of interest `mpg` and `origin`. (If you are working with larger data sets, removing superfluous columns can be an advantage.)
3. We create a new variable `mpg_gt_gaverage` which finds out whether an automobile has an MPG greater than the global average. 
4. We filter all rows to only include American cars. 
5. We <mark>summarised</mark> the data calculate, which helped us find that only 27 percent of American-made cars had MPGs greater than the global average.

::: {.callout-tip collapse="false"}
Think of the pipe as "take this data and do something with it"
:::

::: {.callout-tip collapse="true"}
## How would I do this in base R?

Let's look at how to recreate this using base R.

```{r}
Auto_cleaned <- Auto[,c('mpg','origin')]

Auto_cleaned$mpg_gt_gavg <- if_else(Auto_cleaned$mpg > mean(Auto_cleaned$mpg), TRUE, FALSE) 

Auto_cleaned <- Auto_cleaned[Auto_cleaned$origin == 1, ]

data.frame(prop_mpg_gt_gavg = mean(Auto_cleaned$mpg_gt_gavg))
```

While this code is technically correct, notice a few things. We need to keep updating the same object to save our results. Our code is disjointed and difficult to understand. The final product is also less satisfactory: we needed to convert it from a vector to a data frame, which displays no information on the class of `prop_mpg_gt_gavg`. That is, perhaps, why R can appear so confusing for first-time learners.

:::

## Step 1.4 PRACTICE! (15 min)

Now let's use some of these skills to classify `Chinstrap` penguins. This is data from the package [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/) [^1]. 

```{r}
# Importing this library will make `penguins` data available
library(palmerpenguins) 

head(penguins)
```

🎯 **ACTION POINTS** 

Use the **verbs** you learned above to modify the original data set and create a new one, call it `penguins_cleaned`, according to the following steps:

1. Remove any observations with missing data. 
    💡You can pipe `na.omit()` into you sequence of commands to achieve this. 
2. Next, create a binary variable, call it `chinstrap` that stores TRUE if the penguin is a `Chinstrap`, FALSE otherwise.
3. Now, filter the dataset and keep only the following variables:
    - `is_chinstrap` (our outcome of interest)
    - `bill_length_mm`
    - `bill_depth_mm`

Can you do it without looking at the solution?

::: {.callout-tip collapse="true"}
## Click here to view the solution

```{r}

penguins_cleaned <-
  penguins %>% 
  na.omit() %>% 
  mutate(is_chinstrap = if_else(species == 'Chinstrap', TRUE, FALSE)) %>% 
  select(is_chinstrap, bill_length_mm, bill_depth_mm)

```

:::

If you did it right, you should be able to run the code below and get same plot as shown here in the page:

```{r, dpi=300}

penguins_cleaned %>% 
  ggplot(aes(bill_length_mm, bill_depth_mm, colour=is_chinstrap)) +
  geom_point(size=2.5, stroke=1.4, alpha=0.8, shape=21) +

  # (Optional) customizing the plot
  theme_bw()
```

We can see that Chinstrap penguins tend to have above average bill length and depth whereas the other two species of penguins tend to either have shallow yet long or deep yet short bills.

::: {.callout-important collapse="true"}
## _"What if I struggle to understand the pipe and these "verbs" above?"_

The R for Data Science book is a great resource to learn more about the tidyverse. For more guided tips related to our course, check section [Useful Links about tidyverse](/weeks/week06/checklist.html#useful-links-about-tidyverse) of the ✅ Week 06 - Checklist.

:::

# Step 2: Decision Tree model (50 mins)

Note that the data exhibited in the plot above is not linearly separable. A non-linear algorithm will thus be highly useful in this context. This is what we will explore in this and in the next two sections.

> If you need to recap on non-linear algorithms, rewatch [👨‍🏫 Week 05 - Lecture](/weeks/week05/lecture.html) or check the slides online.

## Step 2.1: Train and visualise a Decision Tree (10 mins)

Since we know the data we are modelling has a nonlinear relationship, let's train a **Decision Tree** to classify Chinstrap penguins. R does not come with Decision Trees installed, so we need to import it from a library. Here we will use the function `rpart` from the [`rpart` package](https://cran.r-project.org/web/packages/rpart/vignettes/longintro.pdf) and `rpart.plot` from the [`rpart.plot` package](http://www.milbo.org/rpart-plot/index.html).

Here are a few things to know about the `rpart` function:

- The `rpart` command is largely similar to other commands such as `lm` and `glm` in that the first parameter is a formula and the second is the data set. 
- We have to add `method = class` to tell the algorithm that we are performing a classification task. 

```{r, dpi=300}

tree.model <- rpart(is_chinstrap ~ ., data = penguins_cleaned, method = 'class')

rpart.plot(tree.model)
```

🤝 **WORKING TOGETHER** In pairs, discuss what you see in the plot.

- What do each node represents?
- What are the numbers inside the nodes?
- What do the percentages represent?

**The video below explains how to read the numbers inside the nodes of the decision tree**

<div style="position: relative; padding-bottom: 47.708333333333336%; height: 0;"><iframe src="https://www.loom.com/embed/76d898bbb7e64d28a041497b47a14ea7" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>

**Alternative visualisation with `parttree` package**

Alternatively, you could visualise a decision tree as a partition tree plot. For this, you will need to have the `parttree` package installed (check instructions at the top of the page).

```{r}

ggplot() +
  geom_point(data = penguins_cleaned, 
             aes(bill_length_mm, bill_depth_mm, colour = is_chinstrap)) +
  geom_parttree(data = tree.model, aes(fill = factor(is_chinstrap)), alpha = 0.25) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom') +
  scale_colour_lancet() +
  scale_fill_lancet() +
  labs(x = 'Bill length (mm)', y = 'Bill depth (mm)',
       colour = 'Is actually chinstrap?', 
       fill = 'Is predicted chinstrap?')

```

_Don't understand what the code above does? You might want to read [Chapter 3 of R for Data Science](https://r4ds.had.co.nz/data-visualisation.html) to review `ggplot`._


## Step 2.2: Goodness-of-Fit of the Decision Tree (10 mins)

Let's investigate how well our model fits the data. Let's reuse the model we trained (`tree.model`) and predict the _same samples_ we used to train it. To avoid modifying our original dataframe, let's save the output of the prediction in an auxiliary df (`plot_df`):

```{r, output=FALSE}
plot_df <- 
    penguins_cleaned %>% 
    mutate(class_pred = predict(tree.model, newdata = ., type="class"),
           correct    = class_pred == is_chinstrap)
plot_df
```

🤝 **WORKING TOGETHER** In pairs, discuss the following:

- Explain what you see in the output of the chunk of code above.
- What does the code above do?

Simple confusion matrix:

```{r, output=FALSE}
confusion_matrix <- 
    table(expected=plot_df$is_chinstrap, class_pred=plot_df$class_pred)
print(confusion_matrix)
```

Nicer looking confusion matrix:

```{r, output=FALSE}
library(cvms)

plot_confusion_matrix(as_tibble(confusion_matrix), 
                      target_col = "expected", 
                      prediction_col = "class_pred",
                      
                      # Customizing the plot
                      add_normalized = TRUE,
                      add_col_percentages = FALSE,
                      add_row_percentages = FALSE,
                      counts_col = "n",
                      )
```

## Step 2.3 Control the parameters! (15 min)

We can tweak how the tree is built by controlling for certain parameters of the algorithm. To take a look at all possible parameters you can control, open the R console, type the command below and hit ENTER:

```r
?rpart.control
```

For example, let's reduce the `minbucket` parameter:

```{r}
tree.model <- rpart(is_chinstrap ~ ., data = penguins_cleaned, method = 'class', control=list(minbucket=1))

rpart.plot(tree.model)
```

🤝 **WORKING TOGETHER** In pairs, discuss the following:

- Explain what is different in this model.
- Visualize the new tree using the `parttree` package (reuse from Step 2.1) 

## Step 2.4 PRACTICE! (15 min)

🎯 **ACTION POINTS**

- Build new tree models, this time trying out different control parameters.
- Which parameters led to different trees? 
- Which parameters change the tree the most?

----------------

🔜 Next week, we will continue our journey of supervised learning by exploring the Support Vector Machine algorithm. We will also visit once again the topic of resampling (cross-validation) to select the optimal parameters for classifiers and regressors.

# 🏠 Take-home exercises

The following exercises were designed as natural extension of the concepts you learned about in this lab session. Reserve some time later to solve this at home, we will release solutions at the end of the week after everyone has had a chance to attend a lab session.

## Q1: Selecting columns

**Using the `Bikeshare` data set present in `ISLR2` subset the data to only include `mnth`, `holiday`, and `bikers` columns.**

```{r}
# Your code goes here
```

## Q2: A new tidyverse verb

**Using the `Bikeshare` data set present in `ISLR2`, replicate Step 1.3, only this time replace `filter(origin == 1)` with `group_by(origin)`. How have the results changed?**

```{r}
# Your code goes here
```

> Your text goes here

## Q3: Exploratory Data Analysis (Part I)

**Calculate the average daily number of bikers in March, in the `Bikeshare` data set.**

```{r}
# Your code goes here
```

## Q4: Exploratory Data Analysis (Part II0

**Do people bike more during holiday seasons?**

```{r}
# Your code goes here
```

> Your text goes here

## Q5: Back to penguins...

**Let's go back to the `penguins` data set. Create a new dataframe that omits missing values (NAs), remove the `island` and `year` columns but keep the rest of the dataset intact (don't create new columns).**

```{r}
# Your code goes here
```

## Q6. Predict Penguin species

**Build a Decision Tree model to predict the `species` of penguins. Note that the outcome will not be a binary variable this time.**

```{r}
# Your code goes here
```

## Q7. Control parameters

**The decision tree algorithm we are using sets the cost complexity parameter (`cp`) by default as `control = list(cp = 0.01)`. Build a new model with a smaller `cp` value, say `control = list(cp = 0.001)`. Does this increase or reduce the complexity of the tree?**

```{r}
# Your code goes here
```

> Your text goes here


[^1]: Data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station](https://pallter.marine.rutgers.edu/), ]Antarctica LTER](https://pallter.marine.rutgers.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/).