---
subtitle: "DS202 Data Science for Social Scientists"
title: "🗓️ Week 05:<br/>Non-Linear Algorithms - Part II"
author: Dr. [Jon Cardoso-Silva](https://www.lse.ac.uk/DSI/People/Jonathan-Cardoso-Silva) and Dr. [Stuart Bramwell](https://www.lse.ac.uk/Methodology/People/Academic-Staff/Stuart-Bramwell/Stuart-Bramwell)
institute: '[\@LSEDataScience](https://twitter.com/lsedatascience)'
date: 14 October 2022
date-meta: 14 October 2022
toc: true
toc-depth: 1
toc-title: "What we will cover today"
center-title-slide: false
tbl-cap-location: bottom
from: markdown+emoji
format:
  revealjs: 
    html-q-tag: true
    pdf-separate-fragments: true
    code-fold: false
    theme: simple
    slide-number: true
    mouse-wheel: true
    chalkboard: 
      buttons: true
    preview-links: auto
    logo: /figures/logos/LSE Data Science Institute.png
    css: /css/styles_slides.css
    footer: 'DS202 - Data Science for Social Scientists :robot: :juggling_person:'
---

## Support Vector Machines for Classification

- Considered one of the best out-of-the-box classifiers (ISLR)
- Able to accommodate non-linear decision boundaries

## Building Intuition: the Hyperplane

```{r}
library(tidyverse)

sim.data <- crossing(x1 = seq(-1.5, 1.5, 0.1),
                     x2 = seq(-1.5, 1.5, 0.1)) %>%
            mutate(hp = if_else(1 + 2*x1 + 3*x2 > 0, 'Yes', 'No')) 

sim.line <- tibble(x1 = c(-1.5, 1.5),
                   slope = (-1 - 2*x1)/3)

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'bottom',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('#709AE1FF','#FED439FF')) +
    labs(x = 'First variable', y = 'Second variable',
         colour = 'Above hyperplane?')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 < 0$

```{r}

library(ggsci)

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    # geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'none',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('#709AE1FF','white')) +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 > 0$

```{r}

library(ggsci)

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    # geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'none',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('white','#FED439FF')) +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 = 0$

```{r}

library(ggsci)

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2, colour = 'white') +
    geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'none') +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Maximal Marginal Classifier

The linearly separable case

```{r}

library(e1071)

set.seed(123)
x1 <- c(rnorm(50, 1, 1) , rnorm(50, 4, 1))
set.seed(234)
x2 <- c(rnorm(50, 1, 1) , rnorm(50, 4, 1))

sim.data <- tibble(x1 = x1, x2 = x2,
                   cl = as.factor(c(rep('1', 50), rep('2', 50))))
  
sim.data %>%
  ggplot() +
  geom_point(aes(x1, x2, colour = cl), size = 2) +
  theme_minimal() +
  theme(legend.position = 'none') +
  labs(x = 'First variable', y = 'Second variable')

mmc <- svm(cl ~ x1 + x2, data = sim.data, kernel = 'linear')

sim.data %>% 
  mutate(support_vects = if_else(row_number() %in% mmc$index, 'Yes', 'No')) %>% 
  ggplot(aes(x1, x2, colour = cl, shape = support_vects, size = support_vects)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(1, 4)) +
  scale_size_manual(values = c(1, 10)) +
  theme_minimal() +
  theme(legend.position = 'bottom') +
  labs(x = 'First variable', y = 'Second variable',
       shape = 'Support vectors?', colour = 'Class')

```