---
subtitle: "Non-linear algorithms"
title: "🗓️ Week 05:<br/> Decision Trees"
author: Dr. [Jon Cardoso-Silva](https://www.lse.ac.uk/DSI/People/Jonathan-Cardoso-Silva) and Dr. [Stuart Bramwell](https://www.lse.ac.uk/Methodology/People/Academic-Staff/Stuart-Bramwell/Stuart-Bramwell)
institute: '[\@LSEDataScience](https://twitter.com/lsedatascience)'
date: 28 October 2022
date-meta: 28 October 2022
toc: true
toc-depth: 1
toc-title: "What we will cover today"
center-title-slide: false
from: markdown+emoji
server: shiny
format:
  revealjs: 
    pdf-separate-fragments: true
    code-fold: false
    theme: simple
    slide-number: true
    mouse-wheel: true
    chalkboard: 
      buttons: true
    preview-links: auto
    logo: /figures/logos/LSE Data Science Institute.png
    css: /css/styles_slides.css
    footer: 'DS202 - Data Science for Social Scientists :robot: :juggling_person:'
---

# The limits of classic regression models 

## The limits of classic regression models 

Linear and logistic regression are a good first shot for building ML models

::: incremental
- Easy-to-interpret coefficients
- Intuitive (ish) ways to assess variable importance
- Often good out-of-the-box predictions
:::

## However...

::: incremental
- Assumption that the predictors are linearly related to the outcome is restrictive
- We have seen, for instance, that accounting for higher order polynomial relationships can produce better model fit 
:::

----


### Example

- Think of the relationship between `lstat` and `medv` in `Boston` ([💻 Week 05 Lab](/weeks/week05/lab.html#step-1.3-does-it-get-better-if-i-modify-the-features-using-polynomial-terms))


```{r}
library(tidyverse)

lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = ISLR2::Auto)
sim.data <- data.frame(horsepower = 46:230)

sim.pred <- predict(lm.fit2, sim.data, interval = 'confidence') 

sim.data <- cbind(sim.data, sim.pred)

ggplot(data = sim.data, aes(x = horsepower, y = fit)) +
   geom_point(data = ISLR2::Auto, aes(x = horsepower, y = mpg)) +
   geom_line() +
   geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.125, fill = 'blue') +
   theme_minimal() +
   labs(x = 'Horsepower', y = 'MPG')
```

## Enter non-linear methods

::: incremental
- These algorithms do not make (strong) statistical assumptions about the data
- The focus is more on predictive rather than explanatory power
:::

# The Decision Tree

<iframe id="reddit-embed" src="https://www.redditmedia.com/r/funny/comments/1j4gf7/cats_decision_tree/?ref_source=embed&amp;ref=share&amp;embed=true&amp;" sandbox="allow-scripts allow-same-origin allow-popups" style="border: none;" height="527" width="640" scrolling="no"></iframe>

## Decision Tree for a Regression task {.smaller}

Using the `Auto` dataset, predict `mpg` with a <mark>tree-based model</mark> using `weight` and `year` as features.

::: {style="margin-top:-1em"}
```{r fig.dpi=300}
#| echo: false
#| code-fold: show

library(ISLR2)
library(tidyverse)
library(rpart)
library(rpart.plot)

tree.reg <- rpart(mpg ~ weight + year, data = Auto, control = list(maxdepth = 2))

rpart.plot(tree.reg, ycompact=FALSE)
```
:::

----

### Source Code {.smaller}

::: {.callout-tip}
- Use the code below to replicate the plot from the previous slide. 
- Found a bug? Report it on Slack.
- 💡 Check out [this tutorial](http://www.milbo.org/rpart-plot/prp.pdf) of `rpart.plot`.
:::

```r
library(ISLR2)      # to load Boston data
library(tidyverse)  # to use things like the pipe (%>%), mutate and if_else

library(rpart)      # a library that contains decision tree models
library(rpart.plot) # a library that plots rpart models 

# The function rpart below fits a decision tree to the data
# You can control various aspects of the rpart fit with the parameter `control`
# Type ?rpart.control in the R console to see what else you can change in the algorithm

tree.reg <- rpart(mpg ~ weight + year, data = Auto, control = list(maxdepth = 2))

rpart.plot(tree.reg)
```


## Decision Tree for a Classification task {.smaller}

Using the `Boston` dataset, predict whether `medv` is above the median using `crim` and `tax`:

::: {style="margin-top:-1em"}
```{r fig.dpi=300}
#| echo: false
#| code-fold: show

Boston <- Boston %>%
    mutate(medv_gtmed = if_else(medv > median(medv), TRUE, FALSE))

tree.class <- rpart(medv_gtmed ~ lstat + tax, data = Boston, control = list(maxdepth = 2))

rpart.plot(tree.class)

```
:::

---

### Source Code 

::: {.callout-tip}
- Use the code below to replicate the plot from the previous slide. 
- Found a bug? Report it on Slack.
- 💡 Check out [this tutorial](http://www.milbo.org/rpart-plot/prp.pdf) of `rpart.plot`.
:::

```r
library(ISLR2)      # to load Boston data
library(tidyverse)  # to use things like the pipe (%>%), mutate and if_else

library(rpart)      # a library that contains decision tree models
library(rpart.plot) # a library that plots rpart models 

# Add a column named `medv_gtmed` to indicate whether tax rate is above median
Boston <- Boston %>% mutate(medv_gtmed = if_else(medv > median(medv), TRUE, FALSE))

# The function rpart below fits a decision tree to the data
# You can control various aspects of the rpart fit with the parameter `control`
# Type ?rpart.control in the R console to see what else you can change in the algorithm

tree.class <- rpart(medv_gtmed ~ lstat + tax, data = Boston, control = list(maxdepth = 2))

rpart.plot(tree.class)
```

# How does it work?

## What's going on behind the scenes?

How decision trees work:

::: incremental
- Divide the predictor space into $\mathbf{J}$ distinct regions $R_1$, $R_2$,...,$R_j$.
- Take the mean of the response values in each region

:::

::: {.fragment}
Here's how the regions were created in our regression/classification examples ⏭️
:::

----

### Alternative representation of decision tree (Regression)

```{r}
library(parttree)
library(scales)

set.seed(123)

Auto %>%
   ggplot(aes(x = weight, y = year)) +
   geom_jitter(size = 3, alpha = 0.25) +
   geom_parttree(data = tree.reg, aes(fill = mpg), alpha = 0.2) +
   theme_minimal() +
   theme(panel.grid = element_blank(), legend.position = 'bottom') +
   scale_x_continuous(labels = comma) +
   scale_fill_steps2() +
   labs(x = 'Weight (lbs)', y = 'Year', fill = 'Miles per gallon')
```

----

### Alternative representation of decision tree (Classification)

```{r}
set.seed(123)

Boston %>%
   ggplot(aes(x = lstat, y = tax)) +
   geom_jitter(size = 3, alpha = 0.25) +
   geom_parttree(data = tree.class, aes(fill = medv_gtmed), alpha = 0.2) +
   theme_minimal() +
   theme(panel.grid = element_blank(), legend.position = 'bottom') +
   scale_x_continuous(labels = percent_format(scale = 1)) +
   scale_y_continuous(labels = dollar) +
   scale_fill_steps2() +
   labs(x = 'Proportion lower status', y = 'Tax rate per $10,000', fill = 'Probability above median')

```

----

#### Source code

::: {.callout-tip}
- Use the code in the following slides to replicate the plot from those two plots. 
- Found a bug? Report it on Slack.
- 💡Check out the [`parttree` documentation](http://grantmcdermott.com/parttree/) for how to customize your plot
- 💡Learn more about data visualisation with ggplot2 on [R for Data Science - Chapter 3](https://r4ds.had.co.nz/data-visualisation.html)
:::

----

#### Source Code (regression)

First, you will have to install the `parttree` package:

```r
# Follow the instructions by the developers of the package
# (https://github.com/grantmcdermott/parttree)

install.packages("remotes")
remotes::install_github("grantmcdermott/parttree", force = TRUE)
```

Then:

```r
library(ISLR2)      # to load Boston data
library(tidyverse)  # to use things like the pipe (%>%), mutate and if_else

library(rpart)      # a library that contains decision tree models
library(parttree)   # R package for plotting simple decision tree partitions

# The function rpart below fits a decision tree to the data
# You can control various aspects of the rpart fit with the parameter `control`
# Type ?rpart.control in the R console to see what else you can change in the algorithm

tree.reg <- rpart(mpg ~ weight + year, data = Auto, control = list(maxdepth = 2))

Auto %>%
   ggplot(aes(x = weight, y = year)) +
   geom_jitter(size = 3, alpha = 0.25) +
   geom_parttree(data = tree.reg, aes(fill = mpg), alpha = 0.2) +
   theme_minimal() +
   theme(panel.grid = element_blank(), legend.position = 'bottom') +
   scale_x_continuous(labels = scales::comma) +
   scale_fill_steps2() +
   labs(x = 'Weight (lbs)', y = 'Year', fill = 'Miles per gallon')
```

---

#### Source Code (classification)

First, you will have to install the `parttree` package:

```r
# Follow the instructions by the developers of the package
# (https://github.com/grantmcdermott/parttree)

install.packages("remotes")
remotes::install_github("grantmcdermott/parttree", force = TRUE)
```

Then:

```r
library(ISLR2)      # to load Boston data
library(tidyverse)  # to use things like the pipe (%>%), mutate and if_else

library(rpart)      # a library that contains decision tree models
library(parttree)   # R package for plotting simple decision tree partitions

# Add a column named `medv_gtmed` to indicate whether tax rate is above median
Boston <- Boston %>% mutate(medv_gtmed = if_else(medv > median(medv), TRUE, FALSE))

# The function rpart below fits a decision tree to the data
# You can control various aspects of the rpart fit with the parameter `control`
# Type ?rpart.control in the R console to see what else you can change in the algorithm

tree.class <- rpart(medv_gtmed ~ lstat + tax, data = Boston, control = list(maxdepth = 2))

Boston %>%
   ggplot(aes(x = lstat, y = tax)) +
   geom_jitter(size = 3, alpha = 0.25) +
   geom_parttree(data = tree.class, aes(fill = medv_gtmed), alpha = 0.2) +
   theme_minimal() +
   theme(panel.grid = element_blank(), legend.position = 'bottom') +
   scale_x_continuous(labels = scales::percent_format(scale = 1)) +
   scale_y_continuous(labels = dollar) +
   scale_fill_steps2() +
   labs(x = 'Proportion lower status', y = 'Tax rate per $10,000', fill = 'Probability above median')
```


## How are regions created? {.smaller}

Recursive binary splitting

::: columns
::: {.column style="width: 41%;border-style: dotted;border-width: 1px;border-color: \"#fefefe\"; margin-right:0.5%;padding:1%;"}
### Top down {style="margin: 0em;"}

- Start from the top of the tree
- Then perform splits at a current level of depth 
:::

::: {.column style="width: 41%;border-style: dotted;border-width: 1px;border-color: \"#fefefe\"; margin-right:0.5%;padding:1%;"}
### Greedy {style="margin: 0em;"}

- Splits are "local" not global
- Only cares about data in the current branch

:::
:::

----

### For regression...

::: {.fragment}
- The tree selects a predictor $X_j$ and a cutpoint $s$ that minimises the residual sum of squares.
:::

:::{.fragment}
- We define two half planes $R_1(j,s) = \left\{X|X_j < s\right\}$ and $R_2(j,s) = \left\{X|X_j \geq s\right\}$ and find $j$ and $s$ by minimising.

$$
\sum_{i: x_i \in R_1(j,s)} (y_i - \hat{y}_{R_1})^2 + \sum_{i: x_i \in R_2(j,s)} (y_i - \hat{y}_{R_2})^2 
$$
:::

----

### For classification...

::: incremental
- The tree selects a predictor $X_j$ and a cutpoint $s$ that maximises <mark>node purity</mark>.
- Gini index: $G = \sum_{k = 1}^{K} \hat{p}_{mk}(1 - \hat{p}_{mk})$ 
- Entropy: $D = - \sum_{k = 1}^{K} \hat{p}_{mk}\log \hat{p}_{mk}$
:::

# What can go wrong

## When trees run amock

::: incremental
- Trees can become too complex if we are not careful
- It can lead to something called <mark>overfitting</mark>
   - High training set predictive power
   - Low test set predictive power
- Let's see one example ⏭️
:::

-----

### The following tree is TOO specialised

```{r}

auto_tbl <-
  Auto %>%
  as_tibble() %>%
  select(-name) %>%
  mutate(origin = as.factor(origin))

tree.unpruned <-rpart(mpg ~ weight + year, data = auto_tbl, control = list(cp = 0.00001))

rpart.plot(tree.unpruned)
```

-----

#### Partition visualisation of the same tree

```{r}
auto_tbl %>%
   ggplot(aes(x = weight, y = year)) +
   geom_jitter(size = 3, alpha = 0.25) +
   geom_parttree(data = tree.unpruned, aes(fill = mpg), alpha = 0.2) +
   theme_minimal() +
   theme(panel.grid = element_blank(), legend.position = 'bottom') +
   scale_x_continuous(labels = scales::comma) +
   scale_fill_steps2() +
   labs(x = 'Weight (lbs)', y = 'Year', fill = 'Miles per gallon')
```

# How to fix it

## Pruning the tree {.smaller}

::: incremental
- <mark>Hyperparameters</mark> are model-specific dials that we can tune
   - Things like `max tree depth`, or `min samples per leaf`
- As with model selection, there is no one one-size-fits-all approach to hyperparameter tuning. 
- Instead, we experiment with resampling
   - Most frequently, <mark>k-fold cross-validation</mark>
:::

::: {.fragment}

::: {.callout-note}
## k-fold cross-validation

- We experimented with k-fold CV in [🗓️ Week 04's lecture/workshop](/weeks/week04/lecture.html#the-cross-validation-setup)
- We will revisit this topic in [🗓️ Week 07's lab](/main/syllabus.html)
- Not compulsory for [✍️ Summmative Problem Set (01) | W05-W07](https://lse-dsi.github.io/lse-ds202-course-notes/assessments/summative1.html)

:::

:::

## Cost Complexity {.smaller}

- We apply $\alpha$ which is a non-negative value to prune the tree. 
- For example, when $\alpha = 0.02$ we can create a less complex tree.

::: {style="margin-top:-1em"}

```{r}

tree.pruned <- rpart(mpg ~ ., data = auto_tbl, control = list(cp = 0.02))

rpart.plot(tree.pruned)

```

:::

# What's Next

After our 10-min break ☕:

- Support Vector Machine
- Tips for the **Summative Problem Set 01**