---
subtitle: ""
title: "🗓️ Week 05:<br/>Non-linear algorithms (Support Vector Machines)"
author: Dr. [Jon Cardoso-Silva](https://www.lse.ac.uk/DSI/People/Jonathan-Cardoso-Silva) and Dr. [Stuart Bramwell](https://www.lse.ac.uk/Methodology/People/Academic-Staff/Stuart-Bramwell/Stuart-Bramwell)
institute: '[\@LSEDataScience](https://twitter.com/lsedatascience)'
date: 14 October 2022
date-meta: 14 October 2022
toc: true
toc-depth: 1
toc-title: "What we will cover today"
center-title-slide: false
tbl-cap-location: bottom
from: markdown+emoji
format:
  revealjs: 
    html-q-tag: true
    pdf-separate-fragments: true
    code-fold: false
    theme: simple
    slide-number: true
    mouse-wheel: true
    chalkboard: 
      buttons: true
    preview-links: auto
    logo: /figures/logos/LSE Data Science Institute.png
    css: /css/styles_slides.css
    footer: 'DS202 - Data Science for Social Scientists :robot: :juggling_person:'
---

```{r}

library(e1071)
library(ggsci)
library(tidyverse)

```

## Support Vector Machines for Classification

- Considered one of the best out-of-the-box classifiers (ISLR)
- Able to accommodate non-linear decision boundaries

## Building Intuition: the Hyperplane

```{r}

sim.data <- crossing(x1 = seq(-1.5, 1.5, 0.1),
                     x2 = seq(-1.5, 1.5, 0.1)) %>%
            mutate(hp = if_else(1 + 2*x1 + 3*x2 > 0, 'Yes', 'No')) 

sim.line <- tibble(x1 = c(-1.5, 1.5),
                   slope = (-1 - 2*x1)/3)

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'bottom',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('#709AE1FF','#FED439FF')) +
    labs(x = 'First variable', y = 'Second variable',
         colour = 'Above hyperplane?')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 < 0$

```{r}

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    theme_minimal() +
    theme(legend.position = 'none',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('#709AE1FF','white')) +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 > 0$

```{r}

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2) +
    theme_minimal() +
    theme(legend.position = 'none',
          panel.grid = element_blank()) +
    scale_colour_manual(values = c('white','#FED439FF')) +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Hyperplane (cont.)

When $1 + 2x_1 + 3x_2 = 0$

```{r}

sim.data %>%
    ggplot() +
    geom_point(aes(x1, x2, colour = hp), size = 2, colour = 'white') +
    geom_line(data = sim.line, aes(x = x1, y = slope), size = 1) +
    theme_minimal() +
    theme(legend.position = 'none',
          panel.grid = element_blank()) +
    labs(x = 'First variable', y = 'Second variable')

```

## Building Intuition: the Maximal Marginal Classifier

The linearly separable case. The x's represent the so-called support vectors.

```{r}

set.seed(123)
x1 <- c(rnorm(50, 1, 1) , rnorm(50, 4, 1))
set.seed(234)
x2 <- c(rnorm(50, 1, 1) , rnorm(50, 4, 1))

sim.data <- tibble(x1 = x1, x2 = x2,
                   cl = as.factor(c(rep('1', 50), rep('2', 50))))
  
sim.data %>%
  ggplot() +
  geom_point(aes(x1, x2, colour = cl), size = 2) +
  scale_colour_lancet() +
  theme_minimal() +
  theme(legend.position = 'none') +
  labs(x = 'First variable', y = 'Second variable')

```

## Identifying support vectors

```{r}

mmc <- svm(cl ~ x1 + x2, data = sim.data, kernel = 'linear')

sim.data %>% 
  mutate(support_vects = if_else(row_number() %in% mmc$index, 'sv', 'o')) %>% 
  ggplot(aes(x1, x2, colour = cl, label = support_vects)) +
  geom_text(size = 5) +
  scale_shape_manual(values = c(1, 4)) +
  scale_size_manual(values = c(1, 10)) +
  scale_colour_lancet() +
  theme_minimal() +
  theme(legend.position = 'bottom',
        plot.caption = element_text(hjust = 0.5, vjust = -1, size = 12.5)) +
  labs(x = 'First variable', y = 'Second variable',
       label = 'Support vectors?', colour = 'Class',
       caption = 'sv = support vector')

```

## When data is not linearly separable

Suppose we have a case that is not linearly separable like this. We have two classes but class 1 is "sandwiched" in between class 2.

```{r}

set.seed(1)
x <- matrix(rnorm (200 * 2), ncol = 2)
x[1:100, ] <- x[1:100, ] + 2
x[101:150, ] <- x[101:150, ] - 2
y <- c(rep(1, 150) , rep(2, 50))
sim.train <- data.frame(x1 = x[,1], x2 = x[,2], y = as.factor(y))

sim.train %>%   
  ggplot() +
  geom_point(aes(x1, x2, colour = y), size = 3) +
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'First variable', y = 'Second variable',
       colour = 'Class')

```

## Enter support vector machines

Let's start with a (linear) support vector classifier function

$$

f(x) = \beta_0 + \sum_{i\in\mathcal{S}}^{} \alpha_i\left\langle x_i, x_{i'} \right\rangle

$$

A couple of points:

- $\left\langle x_i, x_{i'} \right\rangle$ is the inner product of two observations
- $\alpha_i$ is a parameter fitted for all pairs of training observations
- $i\in\mathcal{S}$ indicates observations that are support vectors (all other observations are ignored by setting all $\alpha_i \notin \mathcal{S}$ to zero.)

## Enter support vector machines (cont.)

We can replace $\left\langle x_i, x_{i'} \right\rangle$ with a \textit{generalisation} of the form $K(x_i, x_{i'})$ where $K$ is a kernel.

- Two well-known kernels
- Polynomial $K(x_i, x_{i'}) = (1 + \sum_{j = 1}^{p} x_{ij}x_{i'j})^d $ where $d > 1$
- Radial or "Gaussian" $K(x_i, x_{i'}) = \text{exp}(-\gamma\sum_{j=1}^{p}(x_{ij}-x_{i'j})^2)$ where $\gamma$ is a positive constant.

## Overfitting illustrated

Simple models can sometimes be better than complex models.

```{r}

svm.model.simple <- 
  svm(y ~ ., data = sim.train, kernel = 'radial')

svm.model.complex <-
  svm(y ~ ., data = sim.train, kernel = 'radial',
      gamma = 10^2,  
      cost = 10^4)

kernel.points <-
  crossing(x1 = seq(-5, 5, 0.1),
           x2 = seq(-5, 5, 0.1)) %>% 
  mutate(class.simple = predict(svm.model.simple, .),
         class.complex = predict(svm.model.complex, .))

sim.train <-
  sim.train %>% 
  as_tibble() %>% 
  mutate(pred.simple = predict(svm.model.simple, .),
         pred.complex = predict(svm.model.complex, .),
         correct.simple = if_else(pred.simple == y, TRUE, FALSE),
         correct.complex = if_else(pred.complex == y, TRUE, FALSE))
```

## Simple model on training set

```{r}
sim.train %>%   
  ggplot() +
  geom_tile(data = kernel.points, aes(x1, x2, fill = class.simple), alpha = 0.325) +
  geom_point(aes(x1, x2, colour = y, shape = correct.simple), size = 3) +
  scale_shape_manual(values = c(4, 1)) +
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'First variable', y = 'Second variable',
       fill = 'Class', colour = 'Class',
       shape = 'Correct prediction?',
       title = paste('Training accuracy =', mean(sim.train$correct.simple)))
```

## Complex model on training set

```{r}
sim.train %>%   
  ggplot() +
  geom_tile(data = kernel.points, aes(x1, x2, fill = class.complex), alpha = 0.325) +
  geom_point(aes(x1, x2, colour = y, shape = correct.complex), size = 3) +
  scale_shape_manual(values = c(1, 4)) +
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'First variable', y = 'Second variable',
       fill = 'Class', colour = 'Class',
       shape = 'Correct prediction?',
       title = paste('Training accuracy =', mean(sim.train$correct.complex)))

```

## Now let's create some testing data

Look what happens when I set a different seed (nothing else changes) to construct a test set.

```{r}

set.seed(2)
x <- matrix(rnorm (200 * 2), ncol = 2)
x[1:100, ] <- x[1:100, ] + 2
x[101:150, ] <- x[101:150, ] - 2
y <- c(rep(1, 150) , rep(2, 50))
sim.test <- data.frame(x1 = x[,1], x2 = x[,2], y = as.factor(y))

sim.test <-
  sim.test %>% 
  as_tibble() %>% 
  mutate(pred.simple = predict(svm.model.simple, .),
         pred.complex = predict(svm.model.complex, .),
         correct.simple = if_else(pred.simple == y, TRUE, FALSE),
         correct.complex = if_else(pred.complex == y, TRUE, FALSE))

```

## Simple model on test set

```{r}

sim.test %>%   
  ggplot() +
  geom_tile(data = kernel.points, aes(x1, x2, fill = class.simple), alpha = 0.325) +
  geom_point(aes(x1, x2, colour = y, shape = correct.simple), size = 3) +
  scale_shape_manual(values = c(4, 1)) +
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'First variable', y = 'Second variable',
       fill = 'Class', colour = 'Class',
       shape = 'Correct prediction?',
       title = paste('Testing accuracy =', mean(sim.test$correct.simple)))
```

## Complex model on test set

```{r}
sim.test %>%   
  ggplot() +
  geom_tile(data = kernel.points, aes(x1, x2, fill = class.complex), alpha = 0.325) +
  geom_point(aes(x1, x2, colour = y, shape = correct.complex), size = 3) +
  scale_shape_manual(values = c(4, 1)) +
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = 'bottom',
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'First variable', y = 'Second variable',
       fill = 'Class', colour = 'Class',
       shape = 'Correct prediction?',
       title = paste('Testing accuracy =', mean(sim.test$correct.complex)))

```