---
title: "ðŸ’» Week 09 - Lab Roadmap (90 min)"
date: 20 November 2022
date-meta: 20 November 2022
output: html_document
author: Dr. Jon Cardoso-Silva
---

# Setup

:bulb: *Some of you have mentioned that your R version cannot handle tidymodels. We recommend you update R to version 4.2.0 or above.*

## Packages you will need

```{r, message=FALSE, warning=FALSE}
library('tidyverse')   # to use things like the pipe (%>%)
library('tidymodels')  # for model tuning, cross-validation etc.

# Vanity packages:
library('GGally')      # for pretty correlation plot
```

# Step 1: The Data (50 min)

In this lab we will use the following dataset:

> Paine, Jack, 2019, "Replication Data for: Democratic Contradictions in European Settler Colonies", <https://doi.org/10.7910/DVN/LU8IDT>, Harvard Dataverse, V1, UNF:6:d6rzxrgBZEM6Sv5y7PIHpg== \[fileUNF\]

From the academic article:

> Paine, Jack. **"Democratic Contradictions in European Settler Colonies."** World Politics 71, no. 3 (2019): 542--85. [doi:10.1017/S0043887119000029](https://doi.org/10.1017/S0043887119000029).

This panel data shows the evolution of the independence and democracy in European colonies from the year 1600 to 2000.

## Step 1.1. Explore the data

### Load the Data

Download the dataset from Moodle and place it in the same directory of this RMarkdown file. Run the cell below to read it as a data frame:

```{r}
df <- read_csv("colonial_panel.csv")
```

### How many rows and columns are there in the dataset?

```{r}
dim(df)
```

### What are the columns?

```{r}
colnames(df)
```

### Select columns of interest

We will focus on a few columns and use `distinct()` to drop duplicated rows:

```{r}
df_filtered <- 
  df %>% 
  select(cname, year_colonized, year_indep, first_elect_year) %>% 
  distinct(cname, year_colonized, year_indep, first_elect_year, .keep_all = TRUE)
```

:bulb:: The column `first_elect_year` represents the first elected representative body in British territories colonized before 1800.

:dart:: **ACTION POINT**: Type `View(df_filtered)` in the R Console to visualize our data of interest. Locate the **missing values** present in this data frame. What do they mean? How would deal with those?

> your text here

## Step 1.2. Let's create `ggpairs` plots

:handshake: **WORKING TOGETHER.** We will split the class into three groups. Each group will have to explain one row of the pairs plot.

-   Group 01: Describe the first row
-   Group 02: Describe the second row
-   Group 03: Describe the third row

```{r}
ggpairs(df_filtered %>% select(-cname), progress=FALSE) + theme_bw()
```

> your text here

:speaking_head:: **CLASSROOM DISCUSSIONS**:

-   What are our key takeaways of the plot above?

-   When we run the code above, we see a few messages about *warnings*. What do they tell us?

> your text here

## Step 1.3. Plot `year_indep` x `year_colonized`

```{r}
plot_df <- df_filtered %>% select(cname, year_indep, year_colonized) %>% na.omit()

g <- (
  ggplot(plot_df, aes(x=year_indep, y=year_colonized)) 
  
  + geom_point(alpha=0.4, size=3) 
  
  + theme_bw()
  + labs(x="Year of independence", y="Year colonized", 
         title="Do you spot any clusters?")
)

g
```

**Let's highlight an area of the chart**

We use the [function annotate](https://ggplot2.tidyverse.org/reference/annotate.html) from ggplot and define the boundaries (xmin, xmax, ymin, ymax) on top of the plot we created above

```{r}
g + annotate("rect", alpha=0.1,
             color="red", fill="red", 
             xmin=1930, xmax=2000, 
             ymin=1600, ymax=1700)
```

**In case you want to ZOOM IN :mag_right:...**

You can use `limits=...` of the `scale_x_continuous` and `scale_y_continuous` to control the limits of the X & Y axis of the plot, plus `geom_text` to plot the name of the countries instead of the circles:

```{r}

g2 <- (
  ggplot(df_filtered, aes(x=year_indep, y=year_colonized, label=cname)) 
  
  + geom_label(size=3) 
  
  + theme_bw()
  + labs(x="Year of independence", y="Year colonized")
  + scale_y_continuous(breaks=seq(1600, 1680, 10), limits=c(1600, 1680))
  + scale_x_continuous(breaks=seq(1930, 2000, 5), limits=c(1930, 2000))
  + annotate("rect", color="red", alpha=0,
             xmin=1930, xmax=2000, ymin=1600, ymax=1680)
)

g2
```

:dart:: **ACTION POINT**: Reading the plot. When was Bahamas colonized and when did it get independent?

> your text here

## Step 1.4. Find your own clusters!

:handshake: **WORKING TOGETHER.** In groups or in pairs, you are now asked to manually cluster the data in the original graph `g`. Use what you just learned above about `annotate` and add **several rectangles** to the graph `g`. Give each rectangle a unique colour[^1].

[^1]: Take a look at the following website for more creative colours: https://www.color-hex.com/

``` r
# This is just a template of three rectangles
(g 
  + annotate(geom="rect", alpha=0.1, fill="...", colour="...", 
             xmin=..., xmax=..., ymin=..., ymax=...)
  + annotate(geom="rect", alpha=0.1, fill="...", colour="...", 
             xmin=..., xmax=..., ymin=..., ymax=...)
  + annotate(geom="rect", alpha=0.1, fill="...", colour="...", 
             xmin=..., xmax=..., ymin=..., ymax=...)
)
```

:speaking_head:: **CLASSROOM DISCUSSIONS**:

-   Compare how many rectangles other people used in their own plots.

-   How are your rectangles/clusters different to others?

# Step 2. K-Means Clustering (40 min)

As we learned in [:spiral_calendar: Week 07](https://lse-dsi.github.io/lse-ds202-course-notes/weeks/week07/lecture.html#part-ii---unsupervised-learning-clustering-45-50-min), K-means is a technique that identifies clusters automatically. You have to inform how many clusters you want it to find ($k$ clusters) and then the algorithm will do the following:

1.  Initialize $k$ random **centroids**, then proceed to to step 2 below.

2.  Go over each data point and identify the *nearest* centroid. All points associated with a centroid are said to belong to the same cluster. Proceed to step 3 below.

3.  Now, for each cluster, ignore the centroid and calculate the point that corresponds to the **mean** value of each dimension of the data. Think of it as the "centre of gravity" of the cluster. This point is now the new centroid of the cluster. Proceed to step 4 below.

4.  Go over each data point again, and identify the *nearest* centroid. If **all** data points still belong to the same cluster, stop. Otherwise, go back to step 3 above.

*The algorithm can also stop if you loop over Steps 3 and 4 too many times. There is a parameter that controls the maximum number of iterations called **iter.max***.

It is easy to visualize the trajectory of the centroids If we only have two dimensions. This is what we will do here.

## Step 2.1. Initial k centroids

Let's run the k-means but only for **ONE** iteration. That is, after k-means adjusted the random centroids for the first time.

```{r warning=FALSE}
# Say we believe there are three clusters
k <- 10
set.seed(10)

model <- kmeans(plot_df %>% select(year_indep, year_colonized), 
                centers=k, 
                iter.max=1)
model
```

:dart:: **ACTION POINT**: Run the chunk of code below. What do you think each row represents?

> your text here

```{r}
model$centers
```

**Visualize the centroids**

```{r}
df_centroids <- as_tibble(model$centers)
df_centroids$cluster_id <- factor(seq(1, nrow(df_centroids)))

g3 <- (
  ggplot(plot_df, aes(x=year_indep, y=year_colonized))
  
  + geom_text(aes(label=cname), alpha=0.2, size=3) 
  
  + geom_point(data=df_centroids, aes(color=cluster_id), size=4, shape="X") 
  
  + theme_bw()
  + labs(x="Year of independence", y="Year colonized", 
         title="Visualize the k initial centroids")
)

g3
```

:speaking_head:: **CLASSROOM DISCUSSIONS**:

-   Locate India, Cuba and Egypt in the plot above. Considering the centroids, which clusters do you expect will belong to?

## Step 2.2 Colour data points according to their clusters

Using tidymodels, we can include the clusters in our data frame by augmenting it:

```{r}
plot_df <- augment(model, plot_df)
tail(plot_df)
```

And we can use the new column `.cluster` to colour the dots:

```{r}
g4 <- (
  ggplot(plot_df, aes(x=year_indep, y=year_colonized, colour=.cluster)) 
  
  + geom_point(alpha=0.1, size=3)
  # Or use geom_text
  #+ geom_text(aes(label=cname), alpha=0.25, size=3) 
  
  + geom_point(data=df_centroids, aes(color=cluster_id), size=4, shape="X")
  
  + theme_bw()
  + labs(x="Year of independence", y="Year colonized")
)

g4
```

## Step 2.3. Run K-Means fully

Now, let's allow k-means to run fully, not just for one single iteration.

```{r}
k <- 10
set.seed(10)

model_full <- kmeans(plot_df %>% select(year_indep, year_colonized), 
                centers=k, trace=TRUE)
plot_df <- augment(model_full, plot_df)

df_centroids <- as_tibble(model_full$centers)
df_centroids$cluster_id <- factor(seq(1, nrow(df_centroids)))

g5 <- (
  ggplot(plot_df, aes(x=year_indep, y=year_colonized, colour=.cluster)) 
  
  + geom_point(alpha=0.1, size=3)
  # Or use geom_text
  #+ geom_text(aes(label=cname), alpha=0.25, size=3) 
  
  + geom_point(data=df_centroids, aes(color=cluster_id), size=4, shape="X")
  
  + theme_bw()
  + labs(x="Year of independence", y="Year colonized")
)

g5
```

:dart:: **ACTION POINT**: We set the `trace=TRUE` parameter to see what the algorithm is doing. That is why, in addition to the plot, you also got a few messages printed out. What do you think this printed messages tell us?

> your text here

**Now, look at the centroids of this last model**:

```{r}
model_full$centers
```

:dart:: **ACTION POINT**: Compare the centroids above to the ones you got in Step 2.1. How are they similar/different?

> your text here

## Step 2.4 What is the best number of clusters?

Is it k=3, k=10, k=20??? Unfortunately, there is no absolute final objective solution to this question. It usually depends on what you want to do with this data. Do you _want_ to investigate and compare the characteristics of a large number of groups or just a few?

But here we will show you one possible strategy to automate the process of finding an "optimal" number of clusters based on the total **within-cluster sum of squares**, which we are going to refer to as `$tot.withinss`.

(_Your class teacher might need to use the blackboard here_)

**How?**

The code below is a bit tricky so let's go through it step-by-step. 

- We subset the data using `select` and we get rid of the NAs
- We then create a **list column** called `data`. 
  - Tibbles are different from data frames as you can nest an entire object (whether a data frame, model object etc.) into a single cell. 
  - While it may seem like an odd thing to do, look at the next step. 
- Remember `crossing`? We use `crossing` to come up with a new tibble of all possible combinations between `n.clust` which goes from 1 to 20 and `data`.
- Because of the above, we now have the ability to run 20 k-means clustering models using one tibble, and we did not need to write a single for loop! 
- Because we are looping over a list column, we use different varieties of the `map` function (which is part of the `tidyverse` ecosystem).
  - The map which applies a function over the list column, and returns back an object of the same length. 
  - We use `map2` as we want to loop over `data` and `n.clust`. 
- The `kmeans` function takes two parameters to run, the first of which specifies the data, while the second specifies the number of clusters or `centers` to include in the model. 
- Finally, we want to find the total within-cluster sum of squares `tot.withinss`.     - We do this by using `map_dbl` on the `k.means` list column which returns a vector of doubles. 
- To access `tot.withinss`, we use `glance` on `k.means` and include `$tot.withinss`.


```{r}
km.models <-
  df_filtered %>% select(year_colonized, year_indep) %>% na.omit() %>% 
  nest(data = everything()) %>% 
  crossing(n.clust = 1:20) %>% 
  mutate(k.means = map2(.x = data,
                        .y = n.clust, 
                        .f = ~ kmeans(.x, centers = .y)),
         tot.withinss = map_dbl(.x = k.means, .f = ~ glance(.x)$tot.withinss)) 
km.models
```
### Create an elbow plot

Let's create an [elbow plot](https://www.wikiwand.com/en/Elbow_method_(clustering)) to evaluate what number of clusters is most appropriate. 

Generally, we are looking to identify significant elbow points beyond which no significant decreases in TWCSS occur. 

For our data, k=6 or k=10 are probably among the most appropriate. _But do you see how that is a little bit subjective?_

```{r}
km.models %>% 
  ggplot(aes(n.clust, tot.withinss)) +
  geom_line(linetype = 'dashed') +
  geom_point() +
  theme_minimal() +
  theme(panel.grid.minor = element_blank()) +
  scale_x_continuous(breaks = 1:20) +
  scale_y_continuous(labels = comma) +
  labs(x = 'Number of clusters', y = 'Total within cluster sum of squares')
```

# ðŸ  Take-home exercises

## Q1: Cluster of penguins

Load the `palmerpenguins` dataset and filter the data to omit missing values. Ignoring the columns `species` and `year`, run k-means on this data with $k=3$.

How do the clusters identified by k-means compare to the `species` of penguins?

```{r}
# your code here
```

> your text here

## Q2: Elbow plot

Run k-means on the same data, but this time vary k from $\{1, 2, \ldots, 19, 20\}$ and generate an elbow plot. According to the plot, what is a "good" number of clusters?

```{r}
# your code here
```



