---
title: "DS202 2022 MT - Formative Problem Set 01"
author: "DS202 2022MT Teaching Group"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a formative assessment.Formative assessment does not count towards your overall degree classification but is designed to prepare you for the summative (assessed) work that you will complete later in the course.

Although this formative is not graded, we will give you feedback using a "mock mark" (0-100 scale) just so you know how you will be assessed in the next summative assessment, which will be released at the end of Week 05.

**Deadline**: 25 October 2022

**Submit via Moodle**: [DS202 2022 MT - Formative Problem Set 01 link](https://moodle.lse.ac.uk/mod/coursework/view.php?id=1127006)



## 📒 Instructions

**Read carefully**

**1.** Open this `.Rmd` file using RStudio. ^[You can read more about RMarkdown files [here](https://rmarkdown.rstudio.com/).]

**2.** Work on your solutions and fill the allocated spaces with your answers. Each question has one or more empty spaces marked with the text `Your answer goes here`. 

  Text answers are identified as:
  
  > Your text goes here
  
  R code answers are identified as:
  
  ```{r}
  # Your code goes here
  ```
  
  Equation answers are identified as:
  
  $$Your~equation~goes~here$$
  
  You can add more code R code, plots, text or equations to complement your responses if you want to. For extra tips on formatting equations see ^[https://math.meta.stackexchange.com/questions/5020/mathjax-basic-tutorial-and-quick-reference].

**3.** After you answer all the questions, you should press the Knit button, and knit the file to HTML. If you cannot find this button in RStudio, check out [this tutorial](https://www.youtube.com/watch?v=8eBBPVMwTLo). 

**4.** An HTML file will appear in your files. This is what you must submit to Moodle.

**5.** Open [Moodle](https://moodle.lse.ac.uk/mod/coursework/view.php?id=1127006) and submit your HTML file.

⚠️ _If your code has any errors, you might not be able to knit the document. You must fix those errors first._

⚠️ _If you do not know how to answer a solution, leave it empty. Do not add incomplete R scripts as it will lead to knitting errors._

## 🆔 Identification

THIS IS AN ANONYMOUS SUBMISSION. DO NOT INCLUDE YOUR NAME NOR OTHER PERSONAL INFORMATION ANYWHERE IN THIS FILE OTHER THAN YOUR CANDIDATE NUMBER.

```{r}
# Complete the line below by filling it out with your candidate number. 
# If you do not add your candidate number, Knit will throw an error.
candidate_number <- 
```

## ⚙️ Setup

Import required libraries.No need to change if you are not planning to use other packages.

```{r, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(car)
library(ISLR2)
library(tidyverse)
# If you use other packages, add them here
```


## 🎯 Questions

Use the `Carseats` data set in the `ISLR2` package to answer the following questions:

### Q1. Variables

**List the names of the variables in this dataset and whether they are quantitative or qualitative.** (Worth 1/100 mock marks)

> Your text goes here


### Q2. Dimension

**Use R to list the number of rows in the dataset.** (Worth 1/100 mock marks)

```{r}
# Your code goes here
```


### Q3. Visual

**Selecting only the quantitative variables, plot the correlation between variables** (Worth 5/100 mock marks)

_Tip: If you want to use other R packages, add them to the list of libraries at the top._

```{r}
# Your text goes here. 
```

### Q4. Initial Variable Identification

**Based on just an initial inspection of the data, which variables would you select to train an algorithm to predict `Sales`? Why?** (Worth 7/100 mock marks)

> Your text goes here

### Q5. Simple Linear Regression

**Chose ONE SINGLE variable -- any variable -- and fit a linear regression to predict `Sales`. Show the summary of this linear model.** (Worth 3/100 mock marks)

```{r}
# Your code goes here
```

### Q6. Simple Linear Regression - Interpretation

**Provide an interpretation of each coefficient in the model, stating their values, whether they are significant and what they represent. Be careful---some of the variables in the model are qualitative!** (Worth 10/100 mock marks)

> Your text goes here

```{r}
# (optional) Your code goes here
```

### Q7. Simple Linear Regression - Formula

**Write the model in equation form, carefully handling the qualitative variables properly.** (Worth 5/100 mock marks)

$$Your~equation~goes~here$$

### Q8. Multiple Linear Regression

**Chose ONLY THREE variables and fit a linear regression to predict `Sales`. Show the summary of this linear model.** (Worth 3/100 mock marks)

```{r}
# Your code goes here
```

### Q9. Multiple Linear Regression - Interpretation

**Provide an interpretation of each coefficient in the model, stating their values, whether they are significant and what they represent. Be careful---some of the variables in the model are qualitative!** (Worth 10/100 mock marks)

> Your text goes here

```{r}
# (optional) Your code goes here
```

### Q10. Multiple Linear Regression - Formula

**Write the model in equation form, carefully handling the qualitative variables properly.** (Worth 5/100 mock marks)

$$Your~equation~goes~here$$

### Q11. Model comparison

**Which of the two models you created, in questions Q5 and Q8 provide a better fit?** (Worth 10/100 mock marks)

> Your text answer goes here

### Q12. Collinearity

**What is the Variance Inflation Factor (VIF) of each variable?** (Worth 3/100 mock marks)

```{r}
# Your code goes here
```

### Q13. Collinearity (cont.)

**Based on your responses to Q3 and the output of Q12, would you consider ignoring any variables when building a linear model? Why/Why not?** (Worth 7/100 mock marks)

```{r}
# Your code goes here
```

### Q14. Modelling 

**Considering ALL possible combinations of TWO variables in this dataset, find the one linear model that has the smallest Residual Standard Error. Explain how you reached that conclusion, show us the summary of that model and write the model in equation form.** (Worth 15/100 mock marks)

> Your justification goes here

```{r}
# Your code goes here
```

$$Your~equation~goes~here$$

### Q15. Interaction Effects

**Use the `*` and `:` symbols to fit linear regression models with interaction effects. Could you find any model with interactions that fit better than the models you built in Questions Q5 & Q8 & Q14? Feel free to use as many interactions and variables as you prefer. Justify your answer. Explain how you reached that conclusion, show us the summary of that model and write the model in equation form.** (Worth 15/100 mock marks)

> Your justification goes here

```{r}
# Your code goes here
```

$$Your~equation~goes~here$$
