---
title: "DS202 2022MT W06 Drop In session"
author: "Jon"
date: "01/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Drop-In Session

**OBJECTIVE:** Support with R programming skills

DS202: [We will update this notebook together and I'll share it with everyone later.]{style="color:red"}

# Factors (Categorical Variables)

```{r}
# Say we have a vector of categorical strings

sample_vector <- c("blue", "green", "blue", "red")
typeof(sample_vector)
```

```{r}
# Factor is a bult-in R feature
# This is best way of representing categorical variables
factor(sample_vector)
```

```{r}
as.integer(factor(sample_vector))
```

You can force an order:

```{r}
factor(sample_vector, levels=c("red", "green", "blue"))
```

```{r}
as.integer(factor(sample_vector, levels=c("red", "green", "blue")))
```

### Say I want to count the number of "blues"

```{r}
categorical_vector <- factor(sample_vector, levels=c("red", "green", "blue"))
sum(categorical_vector == "blue")
```

## Comparing vectors

### Say I have two string vectors:

-   They have the same length

-   Elements in the same index represents the same "day"

```{r}
vector1 <- c("blue", "red", "green", "green", "blue", "red")
vector2 <- c("red",  "red", "blue", "blue", "blue", "green")
```

### How many blues do they have in common?

```{r}
vector1 == vector2
```

```{r}
vector1 == "blue"
```

Now, put it all together:

```{r}
output <- vector1 == "blue" & vector2 == "blue"
output
```

**Keyword: Logical Operators**

& stands for an AND operation

\| stands for an OR operation

! stands for a NOT operation

Read more about it [here](https://r4ds.had.co.nz/transform.html?q=logical#logical-operators)

```{r}
sum(output) # Count all occurences of TRUE in the vector
```

If I wanted to do it all in a single line:

```{r}
sum(vector1 == "blue" & vector2 == "blue")
```

## Doing the same with dataframes

If I have the same info but now represented as a data frame, how would I count the number of blues in common?

```{r}
# A random dataframe
df <- data.frame(colourA=c("blue", "red", "green", "green", "blue", "red"),
                 colourB=c("red",  "red", "blue", "blue", "blue", "green"))
df

```

You can access each column by using the `$`:

```{r}
df$colourA
```

```{r}
sum(df$colourA == "blue" & df$colourB == "blue")
```

## Now let's imagine we have two dataframes

 

```{r}
# A random dataframe
df1 <- data.frame(observation=c(1, 2, 3, 4, 5, 6),
                  colour=c("blue", "red", "green", "green", "blue", "red"))
df1
```

```{r}
# A random dataframe
df2 <- data.frame(observation=c(1, 2, 3, 4, 5, 5,  6),
                  colour=c("red",  "red", "blue", "blue", "red", "blue", "green"))
df2
```

First, let's calculate whether there was at least one "blue" in each observation.

```{r}
library(tidyverse)

select(df2, observation)
```

**The pipe**

I could do exactly the same thing using the pipe `%>%`

```{r}
df2 %>% select(observation) # no need to pass `df2` to function select
```

```{r}
tail(select(df2,observation),n=1)


tail(
  select(
      df2, 
      observation), 
  n=1)


df2 %>% select(observation) %>% tail(n=1)
```

## Summarise duplicated groupings

Check the idea of `group_by` (a tidyverse feature)

`summarise` and `n()` only works with groupings (`group_by`).

```{r}
# How many colours are there, per observation?
df2 %>% group_by(observation) %>% summarise(count=n())
```

### How many observations of `df2` have at least one colour "blue"?

```{r}
# How many colours are there, per observation?
df2 %>% group_by(observation) %>% summarise(has_blue=any(colour == "blue"))
```

### How many observations of `df1` have at least one colour "blue"?

```{r}
# How many colours are there, per observation?
df1 %>% group_by(observation) %>% summarise(has_blue=any(colour == "blue"))
```

### How do I compare these two new dataframes?

```{r}
# Store those results into variables
df1_summary <- 
    df1 %>% group_by(observation) %>% summarise(has_blue=any(colour == "blue"))

df2_summary <- 
    df2 %>% group_by(observation) %>% summarise(has_blue=any(colour == "blue"))
```

Both dataframes now have the same number of rows, representing the same "observations" and both have a column called `has_blue`. I can compare both like this:

```{r}
sum(df1_summary$has_blue & df2_summary$has_blue)
```

**Keyword: Logical Operators**

& stands for an AND operation

\| stands for an OR operation

! stands for a NOT operation

Read more about it [here](https://r4ds.had.co.nz/transform.html?q=logical#logical-operators)

### Another way to compare two dataframes

Useful if the two dataframes are not aligned

```{r}
output <- merge(df1, df2, by=c("observation", "colour"))
output
```

```{r}
sum(output$colour == "blue")
```

## Remember to save your `mutate`s!

```{r}
# mutate adds a new column to a dataframe
df2 %>% mutate(is_blue=colour == "blue")
```

Note: `mutate` will add a new column but it will NOT update the dataframe. If you want to re-use the new column, you have to save the new dataframe:

```{r}
# df2 does not have a `is_blue` column
df2
```

If I want to updated it to the SAME dataframe, I have to reassign it (using `<-`)

```{r}
df2 <- df2 %>% mutate(is_blue=colour=="blue")
df2
```

## General comparison of dataframes

By manual inspection:

```{r}
table(df1$colour) 
```

```{r}
table(df2$colour)
```

## ggplot2

### - Colour histograms according to a category

I will use `iris` as an example:

```{r}
library(datasets)
data(iris)
iris
```

Generate a histogram of `Petal.Length`:

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length))
  
  + geom_histogram()
  
  # Customize
  + theme_minimal()
)

g
```

How do I colour the histogram according to the `Species`?

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, fill=Species))
  
  + geom_histogram()
  
  # Customize
  + theme_minimal()
)

g
```

Each `geom_` "listens" to a different set of **aesthetics**. (Check [Chapter 3 of R for Data Science](https://r4ds.had.co.nz/data-visualisation.html) for more info)

For example, `geom_point` does not understand the `fill` :

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, y=Petal.Width, fill=Species))
  
  + geom_point()
  
  # Customize
  + theme_minimal()
)

g
```

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, y=Petal.Width, colour=Species))
  
  + geom_point()
  
  # Customize
  + theme_minimal()
)

g
```

### Customizing the colours

#### Manually

```{r}
my_favourite_colours = c("#5bc0de", "#d9534f",  "#ffbf00")
```

Places to find colours: https://www.color-hex.com/color-palettes/popular.php

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, y=Petal.Width, colour=Species))
  
  + geom_point()

  
  # Customize
  + theme_minimal()
  
  + scale_colour_manual(values=my_favourite_colours) # this is where you customize it
)

g
```

#### Built-in palettes

You can use built-in palettes, you just need to know their names/numbers.

-   Check the documentation https://ggplot2.tidyverse.org/reference/scale_brewer.html -\> for the different settings

-   To understand which colour palettes are available, check: https://colorbrewer2.org/

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, y=Petal.Width, colour=Species))
  
  + geom_point()

  
  # Customize
  + theme_minimal()
  
  + scale_colour_brewer(type="qual", palette=2) # Bult-in palette of colours for the aesthetic `colour`, as given by the Colour Brewer
)

g
```

```{r}
df1$source <- "df1"
df2$source <- "df2"
```

```{r}
ggplot(bind_rows(df1, df2), aes(x=colour)) +
  geom_bar() + facet_wrap(~ source)
```

## Faceting

Useful when you want to plot two charts in the same image.

Observation: you might need to combine (append) the two dataframes first. Use the tidyverse function `bind_rows` (same as `rbind`)

```{r}
g <- (
  ggplot(iris, aes(x=Petal.Length, fill=Species))
  
  + geom_histogram()

  
  # Customize
  + theme_bw()
  
  + scale_colour_manual(values=my_favourite_colours) 
  
  + facet_grid(Species ~ .) #This is what does it
)

g
```

## About `tidyverse`

`tidyverse` is a set of R packages that have several functions and facilities for working with data. I find `tidyverse` more intuitive than base R, and there's an entire book available for free online ([R for Data Science](https://r4ds.had.co.nz/index.html)) that contains a lot of helpful tutorials about tidyverse. Let me point to a few specific chapters:

-   You might need or want to transform data when working on your problem sets. Check out [Chapter 5 of R for Data Science](https://r4ds.had.co.nz/transform.html) online book.
-   Data visualization is another helpful skill. You can learn a bit more about ggplot --- the tidyverse way of making plots --- in [Chapter 3 of R for Data Science](https://r4ds.had.co.nz/data-visualisation.html) online book.
-   What should you be looking for when working with data? Check [Chapter 7](https://r4ds.had.co.nz/exploratory-data-analysis.html) to learn the basics of exploratory data analysis.
-   Do you get confused about R Markdown, the idea of "knitting" a file? Then read [Chapter 27](https://r4ds.had.co.nz/r-markdown.html).
-   (More advanced) Do you want to learn how to reshape data or deal with more complex data manipulation? Then have a look at [Chapter 12 - Tidy data](https://r4ds.had.co.nz/tidy-data.html) and [Chapter 13 - Iteration](https://r4ds.had.co.nz/iteration.html).
-   Are you already familiar with tidyverse, but you constantly need to Google how to do things? Save these [cheatsheets](https://www.rstudio.com/resources/cheatsheets/) to your computer.
