---
title: "💻 Week 05 - Lab Roadmap (90 min)"
date: 24 October 2022
date-meta: 24 October 2022
from: markdown+emoji
author: DS202 2022MT Teaching Team
---

Last week we learned how to do the Classification Methods in R. As data scientists, we need to know how to examine the results we generated and also to justify our results.  So, we will focus on and practice the validation methods in the R language in this week's lab. 

If you feel confident at the current stage, free to explore more on your own. We have provided you with some supplementary online resources : 

- 👉 [R for Data Analytics](https://rforanalytics.com/13.6-k-fold-cross-validation.html) (part of the courses by Abhay Singh)
- 👉 [R for Data Science](https://r4ds.had.co.nz/index.html) (book by Hadley Wickham & Garrett Grolemund)

:::{.callout-important collapse="false"}
## Download .Rmd

<mark class="todo"></mark>

Download this roadmap as an .Rmd file <a href="" target="_blank">here</a>.
:::

R packages you will need:

```r
install.packages("tidyverse")
install.packages("ISLR2")
install.packages("rsample")
install.packages("boot")
```


<details><summary> Step 1: The Validation Set Approach (15mins) </summary>

## Step 1: The Validation Set Approach

In this step, we will explore how to extract the subset of the whole dataset as a training dataset, and then estimate the test error rates of various linear models. The dataset is `Auto` from the `ISLR2` package.

### Step 1.1: Training vs Testing splits

Split the `Auto` dataset into two halves, as randomly selecting 196 observations out of the original 392 observations. We performed splits using base R in the last lab. However, this can be done more easily using the `rsample` package. We first create a split object `Auto.split` using `prop = 0.5` to specify a $50\%/50\%$ train/test split. We also specify `strata = mpg` as we want our train/test split to have a similar mean/standard deviation for `mpg`. We then create dataframes using the `training` and `testing` functions.
	
```{r, message=FALSE, warning=FALSE}
library(ISLR2)
library(rsample)

set.seed(1) # Just so you and I have the same "random" numbers.

Auto.split <- initial_split(Auto, prop = 0.5, strata = mpg)
Auto.train <- training(Auto.split)
Auto.test <- testing(Auto.split)
```

:::{.callout-note collapse="true"}
## About `set.seed()`

`set.seed()` is important here as it set a **seed** for the random number generator. Literally, the same results will be replicated in the following steps. Further information can be found in the official documentation.[Seeding Random Variate Generators](https://www.rdocumentation.org/packages/simEd/versions/2.0.0/topics/set.seed) 
:::

### Step 1.2: How good is your model at predicting the test set?

Fit a linear regression model using the training dataset (`Auto.train`), making the `mpg` as the dependent variable(x) and `horsepower` as the independent variable(y). Then, using the fitted model to estimate the `mpg` from `Auto.test`. Finally, calculating the MSE of the 196 observations in the validation set.

```{r, message=FALSE, warning=FALSE}

# use the lm() function to fit a linear regression model
lm.fit <- lm(mpg ~ horsepower, data = Auto.train)

# estimate the 'mpg' values by the lm.fit model
lm.pred <- predict(lm.fit, Auto.test)

# calculate MSE
mean((Auto.test$mpg - lm.pred)^2)

```

Therefore, we have estimated the test MSE for the linear regression model, $\mathbf{25.59}$. (Well Done! 💪 )


:::{.callout-tip collapse="true"}
## How to get help in R

If you are ever in doubt of what a particular R function do, you can consult the R documentation for it by using `?`. 

For example, you have been using `lm` and `predict` but you want to see what other arguments/parameters this function can take, just type `?lm` or `?predict` in your console.
:::


:::{.callout-tip collapse="true"}
## How to access or create columns in R?

In base R, you generally use the `$` sign to access columns or create new columns.

```{r}
#You can see all the values from the 'mpg' column.
Auto$mpg
```
:::


### Step 1.3: Does it get better if I modify the features using polynomial terms?

Repeat the second part to estimate the test error for the quadratic and cubic regressions.

```{r}
lm.fit2 <- lm(mpg ~ poly(horsepower, 2), data = Auto.train)
```

```{r}
mean((Auto.test$mpg - predict(lm.fit2, Auto.test))^2)
```

```{r}
lm.fit3 <- lm(mpg ~ poly(horsepower, 3), data = Auto.train)
mean((Auto.test$mpg - predict(lm.fit3, Auto.test))^2)
```

💡 We can see a model that predicts `mpg` using a quadratic function of `horsepower` performs better than a model that involves only a linear function of `horsepower`. Furthermore, we see that adding a cubic function of `horsepower` actually increases MSE when compared to the quadratic function. Thus, the quadratic function of `horsepower` appears to perform the best out of all the functions considered. 

### Step 1.4: Visualisation

Want to see how well does the quadratic fit maps onto the raw data?

Let's create a scatter plot with `horsepower` on the x-axis and `mpg` on the y-axis. Now, we can predict `mpg` using `lm.fit2` and specify `interval = 'confidence'` to get $95\%$ confidence intervals. Along with `geom_line` we can use `geom_ribbon` to plot the line of best fit and associated confidence intervals. Remember to specify `alpha` so that we can see the predicted value - otherwise the ribbon will not be translucent.

```{r, message = FALSE, warning = FALSE}

    library(tidyverse)

    sim.data <- data.frame(horsepower = 46:230)

    sim.pred <- predict(lm.fit2, sim.data, interval = 'confidence') 

    sim.data <- cbind(sim.data, sim.pred)

    ggplot(data = sim.data, aes(x = horsepower, y = fit)) +
        geom_point(data = Auto, aes(x = horsepower, y = mpg)) +
        geom_line() +
        geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.125, fill = 'blue') +
        theme_minimal() +
        labs(x = 'Horsepower', y = 'MPG')
```

</details>


<details><summary> Step 2: k-Fold Cross-Validation (5mins)</summary>

## Step 2: k-Fold Cross-Validation

It will be easy to follow the former procedure in Step 2, by the `cv.glm` to implement ***K-fold*** CV.

### Step 2.1: Using K=10 folds

Let's estimate Cross-Validation errors corresponding to the polynomial fits of orders one to ten using ten-fold cross-validation (via `K = 10`). 

The `cv.glm()` function is part of the `boot` library. Meanwhile, you can explore the `cv.err` by yourself to see what `call`,`K`,`delta` and `seed` mean. This online webpage will be useful when interpreting the results.

```{r}
library(boot)


set.seed(17)

cv.error.10 <- c() 

for(i in 1:10){
    glm.fit <- glm(mpg ~ poly(horsepower, i), data = Auto)
    cv.error.10[i] <- cv.glm(data = Auto, glmfit = glm.fit, K = 10)$delta[1]
}

cv.error.10
```

```{r}
# we can plot the results by passing a data frame to ggplot

cv.data <- data.frame(poly = 1:10, cv.errs = cv.error.10)

ggplot(data = cv.data, aes(x = poly, y = cv.errs)) +
    geom_point() +
    geom_line(linetype = 'dashed') +
    scale_x_continuous(breaks = 1:10) +
    theme_minimal() +
    labs(x = 'Degree of polynomial', y = 'Cross-validated MSE')
```

Note that in the line of `cv.error.10[i] <- cv.glm(data = Auto, glmfit = glm.fit, K=10)$delta[1]`, it will be very strict to `K` rather than `k`.

</details>

<details><summary> Step 3: The Bootstrap (20mins) </summary>

## Step 3: The Bootstrap

We have learnt the theoretical method regarding Bootstrap. I understand that it may be a bit difficult for beginners in statistics, but we will mainly focus on the coding implementation and visualisation here. Also, we will introduce how to create a `function` below.

**Functions are “self contained” modules of code that accomplish a specific task.** referenced from [Functions and theri argumens](https://hbctraining.github.io/Intro-to-R/lessons/03_introR-functions-and-arguments.html)


With the help of a function, you can reuse the same pattern codes with a simple function name. In fact, you work with functions all the time in R - perhaps without even realising it! 

### Step 3.1 Estimating the Accuracy of a Linear Regression Model

In this step, we will use the boostrap approach to assess the variability of a coefficient estimate. For the sake of simplicity we will look at the relationship between `weight` and `horsepower` which appears to be linear.


Create a function as `boot.fn()` which 

```{r}
boot.fn <- function(data, index) {

    lm(horsepower ~ weight, data = data[index,])$coefficients

}  
```

`boot.fn` simply returns a vector of coefficient estimates. It takes two parameters: `data` and `index`. `data` is a placeholder for the dataframe used in the model. `index` is a placeholder for the sample used to subset the dataframe. Other than this, the body of the function should look familiar. We are estimating a linear model where we are looking to predict `horsepower` by `weight`, and then extracting the coefficients.

Now, compare the results from bootstrap estimates and the standard estimates 
    
```{r}
# bootstrap with 1000 times

boot(Auto, boot.fn, 1000)
```

```{r}
summary(lm(horsepower ~ weight, data = Auto))$coef
```

We can find that in the bootstrap estimation process, $\mathrm{SE}(\hat{\beta}_{0}) = 3.5704$ and $\mathrm{SE}(\hat{\beta}_{1}) = 0.0012$ , while in the standard estimation process, $\hat{\beta}_{Intercept}=-12.1835$ and $\hat{\beta}_{horsepower}=0.0392$. 

To get a better intuition of what the bootstrap algorithm does let's create a ggplot. We can get the intercepts and slopes estimated and overlay them on a scatterplot (`weight` on x-axis, `horsepower` on y-axis). We will create 50 bootstrap resamples for ease of visualisation and use `geom_abline` to overlay all the lines of best fit.

```{r}

boot.model <- boot(Auto, boot.fn, 50)

boot.df <- as.data.frame(boot.model$t)
names(boot.df) <- c('b0','b1')

ggplot(data = Auto, aes(x = weight, y = horsepower)) +
    geom_point() +
    geom_abline(data = boot.df,
                aes(intercept = b0, slope = b1), 
                alpha = 0.1, colour = 'blue') +
    theme_minimal() +
    labs(x = 'Weight (lbs.)', y = 'Engine horsepower')

``` 

</details>

<!-- # 🎯 Practical Exercises -->

<details><summary> Step 4: 🎯 Practical Exercises (50 mins)</summary>

## Step 4: 🎯 Practical Exercises

Since then, we have known and implemented the coding with Cross-validation and Bootstrap.  In this practical case, we will use the new dataset `Default`  and also `Weekly` from the `ISRL` package. **Do not forget to set a random seed before beginning your analysis.**

Some questions are listed below. You are required to try to answer these questions in pairs using R commands. We will go over the solutions once everyone has finished these questions.

### Q1: Train vs test sets

For the `Default` dataset, please split the sample set into a training set and a test set, then fit a logistic regression model that uses `income` and `balance` to predict `default`:

- Q1.1: Use three different splits of the observations into a training set and a test set.
- Q1.2: Fit three multiple logistic regression models using only the training observations.
- Q1.3: Based on the three models, obtain a prediction of default status for each individual in the test set by computing the posterior probability of default for that individual, and classifying the individual to the `default` category if the posterior probability is greater than 0.5.
- Q1.4: Based on the three models, compute the test set error, which is the fraction of the observations in the test set that are misclassified.

### Q2: Bootstrap

For the `Default` dataset, We continue to consider the use of a logistic regression model to predict the probability of `default` using `income` and `balance` on the `Default` data set. 

In particular, we will now compute estimates for the standard errors of the `income` and `balance` logistic regression coefficients in two different ways: 2.1. Using the `bootstrap`, 2.2. Using the standard formula for computing the standard errors in the `glm()` function. As following,

i. Using the `summary()` and `glm()` functions, determine the estimated standard errors for the coefficients associated with `income` and `balance` in a multiple logistic regression model that uses both predictors.


ii. Write a function,`boot.fn()`,that takes as input the `Default` data set as well as an index of the observations, and that outputs the coefficient estimates for `income` and `balance` in the multiple logistic regression model.


iii. Use the `boot()` function together with your `boot.fn()` function to estimate the standard errors of the logistic regression coefficients for `income` and `balance`. Then, Create a histogram of the bootstrap parameter estimates with `ggplot2`, and also set the `bins=20`, title as `1,000 Bootstrap Parameter Estimates - 'balance' & 'income`.
    
</details>






















