---
title: "💻 Week 10 - Lab Roadmap (90 min)"
date: 28 November 2022
date-meta: 28 November 2022
from: markdown+emoji
author: Dr. Jon Cardoso-Silva
---

This week, we will demonstrate how to pre-process a dataset before running any ML model. After preprocessing, we will use Principal Component Analysis (PCA) to explore similarities and dissimilarities in the data.

:::{.callout-important}
Download this page as an RMarkdown file from Moodle.
:::


# Setup

## Packages you will need

```{r, message=FALSE, warning=FALSE}
library(tidyverse)   # to use things like the pipe (%>%)
library(tidymodels)  # for model tuning, cross-validation etc.

# Vanity packages:
library(ggcorrplot)
library(ggsci)       # we like pretty colours
```

# Step 1: The Data (10 min)

In this lab we will use a subset of the [V-Dem v12 dataset](https://www.v-dem.net/data/the-v-dem-dataset/), the world's most comprehensive and detailed democracy ratings. There are 764 variables and indicators in this dataset; you can read about each one of the variables in the [V-Dem v12 Codebook](https://www.v-dem.net/documents/1/codebookv12.pdf). Here we chose just a few columns from the **Historical V-Dem > Elections** data collection (Chapter 4 of the Codebook).

### Load the Data

Let's load the data:

```r
df <- read_csv("VDEM_v12_subset.csv")
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
df <- read_csv("../resources/VDEM_v12_subset.csv")
```

### How many rows and columns are there in the dataset?

```{r}
dim(df)
```

### What are the columns?

```{r}
colnames(df)
```
📚 **REFERENCE**: Here is a summary of the columns we selected:

- `v2eltype_0`: Were there any **first round** legislative elections this `year` in this `country_name`?
- `v2eltype_1`: Were there any **second round** legislative elections this `year` in this `country_name`?
- `v2eltype_6`: Were there any **first round** presidential elections this `year` in this `country_name`? 
- `v2eltype_7`: Were there any **second round** presidential elections this `year` in this `country_name`? 
- `v2asuffrage`: What is the approximate percentage of enfranchised adults older than the minimal voting age?
- `v2elage`: What is the minimum voting age?
- `v2eltrnout`: What is the election turnout?
- `v2elaccept_ord`: Do election losers accept results? From 0 (None) to 4 (All).
- `v2elasmoff_ord`: Following this election, did winners assume office according to prescribed constitutional rules and norms? Range from 0 (No) to 2 (Yes)


### Explore

🎯 **ACTION POINT** Type `View(df)` **on the R console** and hit ENTER to visualise the data.

_`View()` does not work when knitting the file. So, to avoid future problems, as a good practice, it is best to run the `View()` function from the Console and not the R Markdown._

🗣️ **CLASSROOM DISCUSSIONS** Why are there so many NAs in this dataset? Share your observations of this dataset with the class teacher and your colleagues.

# Step 2: Pre-process the data (50 min)

Time to make things a bit more interesting!

Say our goal is to compare countries based on the _frequency_ of national elections (legislative and presidential) as well as things like average minimum voting age or average turnout over a common time period. Also, say we are only interested in data from the year 1900 onwards.

The data we have is not ready for this kind of analysis, so let's spend some time reshaping it. (In real-life data science, 80% of the time is spent doing this kind of data preprocessing!).

## Step 2.1. Add a `period` column

Let's aggregate data about the countries over a common 20-year time window period.

We use the arithmetic operation of [integer division](https://stackoverflow.com/a/66113826/843365) to make this possible here.

```{r}
tmp_df <- 
  df %>%
  filter(year >= 1900, year < 2020) %>%
  mutate(period=paste0("[", 20 * (year %/% 20),
                       "-", 20 * (year %/% 20) + 19, "]"))

# Selecting these columns just to make it easier to visualize
tmp_df %>% 
  select(country_name, year, period) %>%
  head(n=40)
```
## Step 2.2. Count the number of elections per election type

Run the code below and type `View(tmp_df)`: 

```{r}
tmp_df <-
  # The same pre-processing as before
  df %>%
  filter(year >= 1900, year < 2020) %>%
  mutate(period=paste0("[", 20 * (year %/% 20),
                       "-", 20 * (year %/% 20) + 19, "]")) %>%
  
  # This bit is new 
  group_by(period, country_name) %>%
  summarize(across(c(v2eltype_0, v2eltype_1, v2eltype_6, v2eltype_7),
                   ~ sum(., na.rm=TRUE),
                   .names = "sum_{.col}"),
            .groups="keep")
```

🤝 **WORKING TOGETHER:** Now, in groups or in pairs, can you explain why the dimensions our (new) temporary data frame `tmp_df` is different to the original data frame `df`? 

```{r}
dim(tmp_df)
```
🤝 **WORKING TOGETHER:** What do the columns of `tmp_df` represent?

> your notes here


## Step 2.3. Take the average of suffrage levels, voting age and turnout rate

Each `period` can contain multiple elections so variables such as `v2asuffrage`, `v2elage` and `v2eltrnout` cannot be used as they are, we need to _aggregate_ them somehow.

Let's use the same combo of summary+across as before and create columns to contain the _mean_ of these values:

```{r}
tmp_df <-
  # The same pre-processing as before
  df %>%
  filter(year >= 1900, year < 2020) %>%
  mutate(period=paste0("[", 20 * (year %/% 20),
                       "-", 20 * (year %/% 20) + 19, "]")) %>%
  
  # It's a different summarise!
  group_by(period, country_name) %>%
  summarize(across(c(v2asuffrage, v2elage, v2eltrnout),
                   ~ mean(., na.rm=T),
                   .names = "mean_{.col}"),
            .groups="keep")
```

🤝 **WORKING TOGETHER:** In groups or in pairs, take a look at the new `tmp_df`, pay close attention to the NAs in this data frame. Then, answer the following questions: 

- How would you handle these NAs? Should we remove all rows that contain NA or should we replace the NAs by some value (a process called [data imputation](https://www.wikiwand.com/en/Imputation_(statistics)?

> your notes here

🗣️ **CLASSROOM DISCUSSIONS**: Share your thoughts!

## Step 2.4. Winners and Losers

In the original data, `v2elaccept_ord == 4` when the losers of all national elections in a particular year conceded defeat, and `v2elasmoff_ord == 2` when all winners assumed office according to the result of elections without any restrictions.

Out of all elections that happens in a `period`, what percentage of winners fully assumed office and what percentage of losers fully conceded defeat?

```{r}
tmp_df <-
  # The same pre-processing as before
  df %>%
  filter(year >= 1900, year < 2020) %>%
  mutate(period=paste0("[", 20 * (year %/% 20),
                       "-", 20 * (year %/% 20) + 19, "]")) %>%
  group_by(period, country_name) %>%
  
  # This bit is new!
  summarise(pctg_v2elaccept_full=sum(v2elaccept_ord == 4, na.rm=T)/sum(!is.na(v2elaccept_ord)),
            pctg_v2elasmoff_full=sum(v2elasmoff_ord == 2, na.rm=T)/sum(!is.na(v2elasmoff_ord)),
            .groups="keep") 
```

🎯 **ACTION POINT** Once again, type `View(tmp_df)` **on the R console** and hit ENTER to visualise the data.


## Step 2.5 Putting it all together

The code below combines what we have done in Steps 2.1-2.4 above. Run it and `View(df_preprocessed)` to see the output. We will use the `is_all_NAs` column to filter out the true NAs afterwards.

```{r}
df_preprocessed <-
  df %>%
  filter(year >= 1900, year < 2020) %>%
  mutate(period=paste0("[", 20 * (year %/% 20),
                       "-", 20 * (year %/% 20) + 19, "]")) %>%
  
  # It's a different summarise!
  group_by(period, country_name) %>%
  summarize(across(c(v2eltype_0, v2eltype_1, v2eltype_6, v2eltype_7),
                   ~ sum(., na.rm=T),
                   .names = "sum_{.col}"),
            across(c(v2asuffrage, v2elage, v2eltrnout),
                   ~ mean(., na.rm=T),
                   .names = "mean_{.col}"),
            pctg_v2elaccept_full=sum(v2elaccept_ord == 4, na.rm=T)/sum(!is.na(v2elaccept_ord)),
            pctg_v2elasmoff_full=sum(v2elasmoff_ord == 2, na.rm=T)/sum(!is.na(v2elasmoff_ord)),
            .groups="keep") %>% 
    ungroup()
```

How many time periods are we talking about?

```{r}
df_preprocessed %>% distinct(period) %>% as.list()
```

```{r}
dim(df_preprocessed)
```

## Step 2.6. Handling missing data

Where are the NAs?

```{r}
df_preprocessed %>% summarise(across(everything(),~ sum(is.na(.)))) %>% t()
```

Here we choose the "lazy" path, let's just ignore the rows that contain any NA:

```{r}
df_preprocessed <- df_preprocessed %>% na.omit()
```

```{r}
dim(df_preprocessed)
```


💡 You might disagree with this strategy of handling missing data, and that is great! If you have an alternative strategy in mind, do share it with us on Slack! I would be curious to hear about it.


# Step 3. PCA (30 min)

PCA is a **linear** technique that recombines data onto a new set of variables with useful properties. Most commonly, PCA is used to reduce the dimensionality of a dataset and to understand similarities in the data.


## Step 3.1 Training PCA

Here we take a pragmatic approach. Let's try to understand the technique by comparing our current dataset to the new data that will be produced by PCA.

`df_preprocessed` has 11 columns, 9 are predictors whereas the other 2 columns are identifiers (`country_name` and `period`). Let's look at the correlation between the features:


```{r}
df_preprocessed %>% 
  select(-country_name, -period) %>% 
  na.omit() %>% 
  cor() %>% 
  ggcorrplot(outline.col = "white", lab=TRUE, 
             type="lower", lab_size=2.2, tl.cex=10)
```

🏠 **TAKE-HOME ACTIVITY:** This week, instead of coding exercises, we suggest you practice some more `tidymodels`. We will point to a few suggested official tutorials. In this section, we need to use something called `recipes`; read the [Preprocess your data with recipes](https://www.tidymodels.org/start/recipes/) tutorial on tidymodel's website to learn more.

Now, let's see how you could train PCA using the `tidymodels` package:

```{r}
pca_recipe <-
  # First we specify a recipe of what data we are using
  recipe(~ ., data = df_preprocessed) %>%
  
  # Columns period and country_name are not to be used as predictors.
  # That is why we update their roles in the recipe
  update_role(period, new_role = 'id') %>% 
  update_role(country_name, new_role = 'id') %>% 
  
  # PCA requires that data have the same distribution
  # This can be achieved by normalizing the data (mean=0 and std=1)
  step_normalize(all_predictors()) %>% 
  
  # This is where we tell the recipe to run PCA and return 9 Principal Components
  step_pca(all_predictors(), num_comp=9)

# pca_recipe created a recipe, but it didn't run any of those steps.
# To train the PCA, we have to prepare the recipe -- with prep()
pca_prep <- prep(pca_recipe)
```

The object `pca_recipe` contains a recipe and we can choose to prepare a recipe (`prep()`) at any time.

## Step 3.2 Look at the new data

Now that we have prepared our recipe, let's `bake` it using our ingredients (the data):

```{r}
new_df <- bake(pca_prep, df_preprocessed)
```

🎯 **ACTION POINT** Once again, type `View(new_df)` **on the R console** and hit ENTER to visualise the new data frame.

🗣️ **CLASSROOM DISCUSSIONS**: What do you think the columns in this data frame represent?

### How are these variables distributed?

```{r}
plot_df <- pivot_longer(new_df, PC1:PC9)
ggplot(plot_df, aes(x=value, fill=name)) + geom_histogram() + theme_bw() + facet_wrap(~ name)
```
### Correlation plot

🗣️ **CLASSROOM DISCUSSIONS**: Observe the correlation plot below. Why do you think there is no correlation between any predictor?

```{r}
new_df %>% 
  select(-country_name, -period) %>% 
  cor() %>% 
  ggcorrplot(outline.col = "white", lab=TRUE, 
             type="lower", lab_size=2.2, tl.cex=10)
```

💡 _If you create a `ggpairs()` plot, you will see what zero correlation means. Each pairwise combination of features produces a scatter plot that look essentially like a random cloud of points._


## Step 3.3. Looking closer...

These properties of the new data set were produced by design!

> **💡 PCA "rearranges" the original data matrix, producing a new data matrix where all features are intentionally completely uncorrelated to each other.**

This is the key takeaway of PCA. If you remember only one thing about this technique, remember the sentence above!

For example, this is how `PC1` is constructed from the other predictors:

$$
\begin{align}
\text{PC1} = &+ 0.21799702 \times \text{sum_v2eltype_0}^* + 0.24465538 \times \text{sum_v2eltype_1}^* \\
&+0.51687063 \times \text{sum_v2eltype_1}^* + 0.49560420  \times \text{sum_v2eltype_7}^* \\
&+0.44277858 \times \text{mean_v2asuffrage}^* - 0.36757212 \times \text{mean_v2elage}^*\\
&-0.11972783 \times \text{mean_v2eltrnout}^* + 0.07999905 \times \text{pctg_v2elaccept_full}^* \\
&+0.16716920 \times \text{pctg_v2elasmoff_full}^*
\end{align}
$$
where we marked predictors with an asterisk just to remind us that these are not the _raw_ data, but the _normalised_ values of those features.

_The equation above looks familiar, right? Indeed, the way PCA calculates these weights (they are called loadings) bares some similarity to how a linear regression problem is solved._

With the code below, we get to see how PCA calculated the other Principal Components. 

```{r}
pca_prep$steps[[2]]$res
```

What else is there? Let's look at the `summary()` of PCA:

```{r}
summary(pca_prep$steps[[2]]$res)
```

The table below indicates that if we chose to use only the first 5 Principal Components (PC1, PC2, PC3, PC4, PC5) instead of all of them and instead of the original dataframe, we would still have preserved 70% of the variance in this data.

🏠 **TAKE-HOME ACTIVITY:** Check out [this tutorial on Dimensionality Reduction](https://www.tmwr.org/dimensionality.html)