---
title: "🗓️ Week 04 Resampling Workshop"
author: "DS202 2022MT Teaching Group"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

1.  Form groups of \~5 people
2.  

Packages you will need:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ISLR2)
library(cvms)
library(pROC)
library(scales)
```

# Obtain the Data

Follow the links on Slack/Moodle to download

# Exploratory Data Analysis

We are going to work with the `Wage` Data that comes with the `ISLR2` package.

How many rows/columns are there in the dataset?

```{r}
dim(Wage)
```

Let's have a look at it:

```{r}
head(Wage)
```

What are we looking at? Let's understand the distribution of some of these columns:

## Column: `age`

What is the range of values that the column `age` takes?

```{r}
summary(Wage$age)
```

Now look at the distribution:

```{r}
ggplot(Wage, aes(x=age)) + 
  geom_histogram(binwidth=2) +
  
  # Customizing the plot
  theme_bw() +
  scale_x_continuous(breaks=seq(0, 80, 10), limits=c(0, 80)) +
  
  ggtitle("Thankfully, there is no data about child labour!")
```

## Column: `education`

What is the range of values that the column `education` takes?

```{r}
summary(Wage$education)
```

```{r}
ggplot(Wage, aes(x=education)) + 
  geom_bar() +
  
  # Customising the plot
  theme_bw() +
  coord_flip() +

  ggtitle("Most people in this dataset have a '2. HS Grad' level of education")
```

**Take-home Challenge:** Can you convert the plot above to a [donut chart](https://r-graph-gallery.com/128-ring-or-donut-plot.html)?

## Compare `age` vs `education`

```{r, fig.width=9}
ggplot(Wage, aes(x=age, y=education)) +
  
  geom_bin2d() +
  
  # Customising the plot
  theme_bw() +
  scale_fill_viridis_c() +
  
  scale_x_continuous(breaks=seq(0, 80, 10)) +
  
  ggtitle("Comparison age vs education level")
```

## Column: `jobclass`

What is the range of values that the column `jobclass` takes?

```{r}
summary(Wage$jobclass)
```

```{r}
ggplot(Wage, aes(x=jobclass)) + 
  geom_bar() +
  
  # Customising the plot
  theme_bw() +
  coord_flip() +

  ggtitle("Most people in this dataset have a '2. HS Grad' level of education")
```

## What about `wage`?

### Distribution of `wage`:

```{r}
summary(Wage$wage)
```

```{r, message=FALSE, }
ggplot(Wage, aes(x=wage)) + 
  geom_histogram(binwidth=10) +
  
  # Add a line to represent the MEDIAN WAGE  
  geom_vline(xintercept=median(Wage$wage), color="#fb6341", size=1.4, linetype="dashed") + 
  
  # Customising the plot
  theme_bw() +
  
  ggtitle("The median salary is $104k")
```

### Compare `wage` vs `age`:

```{r, fig.width=9, message=FALSE, warning=FALSE}

ggplot(Wage, aes(x=age, y=wage, group=age)) +
  
  geom_boxplot(fill="#008fd4") +
  
  theme_bw() +
  scale_x_continuous(breaks=seq(0, 80, 10)) +
  
  ggtitle("How does wage vary according to age?")

```

### Compare `wage` vs `education`:

```{r, fig.width=7, message=FALSE, warning=FALSE}

ggplot(Wage, aes(x=education, y=wage, group=education)) +
  
  geom_boxplot(fill="#008fd4") +
  
  theme_bw() +
  
  ggtitle("💡 Wage increases with education level")

```

### Compare `wage` vs `jobclass`:

```{r, fig.width=7, message=FALSE, warning=FALSE}

ggplot(Wage, aes(x=jobclass, y=wage, group=jobclass)) +
  
  geom_boxplot(fill="#008fd4") +
  
  theme_bw() +
  
  ggtitle("💡 Information jobs pay slightly larger wages")

```

### Compare `wage` vs `maritl`:

```{r, fig.width=7, message=FALSE, warning=FALSE}

ggplot(Wage, aes(x=maritl, y=wage)) +
  
  geom_boxplot(fill="#008fd4") +
  
  theme_bw() +
  
  ggtitle("How does marital status affect wages?")

```

# Let's convert this to a classification problem

Say we don't care so much about the actual wage but we are interested in predicting whether a particular person makes above $\$ 150k$. Let's create a variable to represent that:

```{r}
new_wage <- data.frame(Wage %>% select(-c(year, logwage))) # Create a copy of Wage
new_wage$above150k <- factor(ifelse(Wage$wage < 150, "No", "Yes"))

cat("How many are there in the sample?\n") # cat is just another way to print something
table(new_wage$above150k)
```

## Separate a sample for external validation

```{r}
library(tidymodels)

external_split <- validation_split(new_wage, prop=0.9)

external_data <- new_wage[-external_split$splits[[1]][[2]],]
table(external_data$above150k)
```

### Split the internal data split into 5 folds

```{r}
internal_data <- new_wage[external_split$splits[[1]][[2]],]
dim(internal_data)
```

```{r}
internal_splits <- vfold_cv(internal_data, v=5)
internal_splits$splits
```

#### Group 01

```{r}

samples_train <- internal_data[internal_splits$splits[[1]][[2]], ]
samples_test  <- internal_data[-internal_splits$splits[[1]][[2]], ]
table(samples_test$above150k)
```

```{r}
saveRDS(samples_train, "Group 01/samples_train.rds")
saveRDS(samples_test, "Group 01/samples_test.rds")
```

#### Group 02

```{r}

samples_train <- internal_data[internal_splits$splits[[2]][[2]], ]
samples_test  <- internal_data[-internal_splits$splits[[2]][[2]], ]
table(samples_test$above150k)
```

```{r}
saveRDS(samples_train, "Group 02/samples_train.rds")
saveRDS(samples_test, "Group 02/samples_test.rds")
```

#### Group 03

```{r}

samples_train <- internal_data[internal_splits$splits[[3]][[2]], ]
samples_test  <- internal_data[-internal_splits$splits[[3]][[2]], ]
table(samples_test$above150k)
```

```{r}
saveRDS(samples_train, "Group 03/samples_train.rds")
saveRDS(samples_test, "Group 03/samples_test.rds")
```

#### Group 04

```{r}

samples_train <- internal_data[internal_splits$splits[[4]][[2]], ]
samples_test  <- internal_data[-internal_splits$splits[[4]][[2]], ]
table(samples_test$above150k)
```

```{r}
saveRDS(samples_train, "Group 04/samples_train.rds")
saveRDS(samples_test, "Group 04/samples_test.rds")
```

#### Group 05

```{r}

samples_train <- internal_data[internal_splits$splits[[5]][[2]], ]
samples_test  <- internal_data[-internal_splits$splits[[5]][[2]], ]
table(samples_test$above150k)
```

```{r}
saveRDS(samples_train, "Group 05/samples_train.rds")
saveRDS(samples_test, "Group 05/samples_test.rds")
```

## Fit a logistic regression

```{r}
model <- glm(above150k ~ age + education + maritl + jobclass, data=new_wage, family=binomial)
#summary(model)
```

### Look at the assigned probabilities for the training data:

```{r, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
target           <- new_wage$above150k
pred_probability <- predict(model, new_wage, type="response")

df_results <- data.frame(target=target, pred_probability=pred_probability)


ggplot(df_results, aes(x=pred_probability, fill=target)) +
  geom_histogram(binwidth=0.05) +
  
  theme_bw() +
  xlim(c(0, 1)) +
  
  ggtitle("This plot compares the predict probability with the target variable (colours)")
```

Where should we cut? What is a **good threshold**?

```{r}
selected_threshold <- 0.50 ## Change this to see the change in the confusion matrix

prediction <- factor(ifelse(pred_probability < selected_threshold, "No", "Yes"))

confusion_matrix <- table(target, prediction)
confusion_matrix
```

Let's make it more visual:

```{r, message=FALSE, warning=FALSE}

library(cvms)


plot_confusion_matrix(as_tibble(confusion_matrix), 
                      target_col = "target", 
                      prediction_col = "prediction",
                      
                      # Customizing the plot
                      add_normalized = TRUE,
                      add_col_percentages = FALSE,
                      add_row_percentages = FALSE,
                      counts_col = "n",
                      )

```

## Calculate some metrics out of the confusion matrix

**ACCURACY:** How many observations did we get right?

```{r}
accuracy <- sum(prediction == target)/nrow(new_wage)
cat(sprintf("%.2f %%", accuracy*100)) # cat is just another way to print() things
```

**SPECIFICITY:** True Negative Rate

```{r}
total_real_no      <- sum(target == "No")
total_correct_no   <- sum(prediction == "No" & target == "No")

TNR <- total_correct_no/total_real_no
cat(sprintf("%.2f %%", TNR*100))

```

**SENSITIVITY:** True Positive Rate

```{r}
total_real_yes      <- sum(target == "Yes")
total_correct_yes   <- sum(prediction == "Yes" & target == "Yes")

TPR <- total_correct_yes/total_real_yes
cat(sprintf("%.2f %%", TPR*100)) 

```

**🎯 EXERCISE 01:** Go back up and tweak the threshold again until you find one that works better for your **training data**!

To help you find out, tweak the parameters below, let us look at the Receiving Operating Characteristic (ROC) curve:

```{r, message=FALSE, warning=FALSE}
library(pROC)

rocobj <- roc(target, pred_probability)


ggroc(rocobj, colour = 'steelblue', size = 2) + 

  xlab("Specificity (TNR)") + 
  ylab("Sensitivity (TPR)") + 

  # THEMING
  theme_bw() +
  scale_x_reverse(labels=scales::percent) +
  scale_y_continuous(labels=scales::percent) +
  
	theme(
        axis.title.x = element_text(size=20,margin = margin(t = 10, r = 0, b = 0, l = 0)),
        axis.title.y = element_text(size=20, margin = margin(t = 0,  r = 10, b = 0, l = 0)),
        legend.text = element_text(size=18),
        legend.position="right",
        axis.text.x = element_text(size=18),
        axis.text.y = element_text(size=18),
        panel.grid.minor = element_blank(),
        aspect.ratio = 1,
        plot.margin=grid::unit(c(0,0,0,0), "mm"))

```
