---
title: "ðŸ—“ï¸ Week 04 Resampling Workshop"
author: "DS202 2022MT Teaching Group"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step 1: Setup

1.  Form groups of \~5 people. Each group will be assigned a specific DATASET by the teacher
2.  Decide who will act as the spokesperson for the group
3.  Check Slack for the link to download the files you will need
4.  Download the RMarkdown from there
5.  Locate the folder that is relevant to your group (ex.: `Dataset 1`) and download the two files from there
6.  Open RStudio, create a new project (File -\> New Project), and move all downloaded files to there
7.  Let's get started

# Step 2: Setup packages

Packages you will need:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(cvms)
library(pROC)
library(scales)
library(e1071)
```

Install any packages you don't have

# Step 3: Check everything is in order

## Step 3.1: Load the data

Run the following. Ensure it doesn't throw an error:

```{r}
train_samples <- readRDS("Group 01/samples_train.rds")
test_samples  <- readRDS("Group 01/samples_test.rds") 
```

## Step 3.2: Check size of datasets

```{r}
dim(train_samples)
```

**This must print: 2160 rows and 10 columns**

```{r}
dim(test_samples)
```

**This must print: 540 rows and 10 columns**

## Step 3.3 Check columns

```{r}
colnames(train_samples)
```

You should see the following

> "age" "maritl" "race" "education" "region" "jobclass" "health" "health_ins" "wage" "above150k"

This data set contains data about people's `wages`. We will take a quick look at the data soon.

Our goal is to predict whether a particular person makes above $\$ 150k$. So, we will be using `above150k` as the independent variable/target for all purposes in this workshop.

## Step 3.4 Check distribution of variable `above150k`

**ðŸŽ¯ ACTION POINT 01:** Run the two cells below then paste the output to the designated space on Slack (if you don't know where to paste it, ask your lecturer for instructions)

```{r}
table(train_samples$above150k)
```

```{r}
table(test_samples$above150k)
```

# Step 4: Exploratory Data Analysis

Let's look at what this data represents:

```{r}
head(train_samples)
```

What are we looking at? Let's understand the distribution of some of these columns:

## Column: `age`

What is the range of values that the column `age` takes?

```{r}
summary(train_samples$age)
```

Now look at the distribution:

```{r}

ggplot(train_samples, aes(x=age, fill=above150k)) + 
  geom_histogram(binwidth=2) +
  
  # Customizing the plot
  theme_bw() +
  scale_x_continuous(breaks=seq(0, 80, 10), limits=c(0, 80)) +
  
  ggtitle("Thankfully, there is no data about child labour!")
```

## Column: `education`

What is the range of values that the column `education` takes?

```{r}
summary(train_samples$education)
```

```{r}
ggplot(train_samples, aes(x=education, fill=above150k)) + 
  geom_bar() +
  
  # Customizing the plot
  theme_bw() +
  coord_flip() +

  ggtitle("Proportionally,\nthere are more people earning above 150k\namongst those with advanced degrees ")
```

**Take-home Challenge:** Can you convert the plot above to a [donut chart](https://r-graph-gallery.com/128-ring-or-donut-plot.html)?

## Column: `jobclass`

What is the range of values that the column `jobclass` takes?

```{r}
summary(train_samples$jobclass)
```

```{r}
ggplot(train_samples, aes(x=jobclass, fill=above150k)) + 
  geom_bar() +
  
  # Customising the plot
  theme_bw() +
  coord_flip() 
```

ðŸ’¡ Feel free to create plots for the other variables if you like.

# Step 5: Fitting a logistic regression model

```{r}
model <- glm(above150k ~ age + education + maritl + jobclass, data=train_samples, family=binomial)
#summary(model)
```

### Look at the assigned probabilities for the training data:

```{r, fig.width=8, fig.height=3, message=FALSE, warning=FALSE}
target_train          <- train_samples$above150k
pred_train_prob       <- predict(model, train_samples, type="response")

df_results <- data.frame(target_train=target_train, pred_train_prob=pred_train_prob)


ggplot(df_results, aes(x=pred_train_prob, fill=target_train)) +
  geom_histogram(binwidth=0.05) +
  
  theme_bw() +
  xlim(c(0, 1)) +
  
  ggtitle("This plot compares the predict probability with the target variable (colours)")
```

Where should we cut? What is a **good threshold**?

```{r}
selected_threshold <- 0.50 ## Change this to see the change in the confusion matrix

pred_train <- factor(ifelse(pred_train_prob < selected_threshold, "No", "Yes"))

confusion_matrix <- table(target_train, pred_train)
confusion_matrix
```

Let's make it more visual:

```{r, message=FALSE, warning=FALSE}

library(cvms)


plot_confusion_matrix(as_tibble(confusion_matrix), 
                      target_col = "target_train", 
                      prediction_col = "pred_train",
                      
                      # Customizing the plot
                      add_normalized = TRUE,
                      add_col_percentages = FALSE,
                      add_row_percentages = FALSE,
                      counts_col = "n",
                      )

```

## Calculate some metrics out of the confusion matrix

**ACCURACY:** How many observations did we get right?

```{r}
accuracy <- sum(pred_train == target_train)/nrow(train_samples)
cat(sprintf("%.2f %%", accuracy*100)) # cat is just another way to print() things
```

**SPECIFICITY:** True Negative Rate

```{r}
total_real_no      <- sum(target_train == "No")
total_correct_no   <- sum(pred_train == "No" & target_train == "No")

TNR <- total_correct_no/total_real_no
cat(sprintf("%.2f %%", TNR*100))

```

**SENSITIVITY:** True Positive Rate

```{r}
total_real_yes      <- sum(target_train == "Yes")
total_correct_yes   <- sum(pred_train == "Yes" & target_train == "Yes")

TPR <- total_correct_yes/total_real_yes
cat(sprintf("%.2f %%", TPR*100)) 

```

**ðŸŽ¯ ACTION POINT 02:** Go back up and tweak the `selected_threshold` until you find one that works "better" for your **training data**!

**ðŸŽ¯ ACTION POINT 03:** Once the previous step is done, paste the following on the designated space on Slack:

``` r
Spokesperson <- "<Replace this by the name of the spokesperson for your group>"
selected_threshold <- <Replace by your selected value>

accuracy <- <Replace by the measure of accuracy you obtained>
TNR <- <Replace by the measure of TNR you obtained>
TPR <- <Replace by the measure of TPR you obtained>
```

## Let's look at the test sample

```{r}
target_test           <- test_samples$above150k
pred_test_prob        <- predict(model, test_samples, type="response")

df_results <- data.frame(target_test=target_test, pred_test_prob=pred_test_prob)


ggplot(df_results, aes(x=pred_test_prob, fill=target_test)) +
  geom_histogram(binwidth=0.05) +
  
  theme_bw() +
  xlim(c(0, 1)) +
  
  ggtitle("This plot compares the predict probability with the target variable (colours)")
```


```{r}
pred_test <- factor(ifelse(pred_test_prob < selected_threshold, "No", "Yes"), levels=c("No", "Yes"))

confusion_matrix <- table(target_test, pred_test)
confusion_matrix
```

Let's make it more visual:

```{r, message=FALSE, warning=FALSE}

library(cvms)


plot_confusion_matrix(as_tibble(confusion_matrix), 
                      target_col = "target_test", 
                      prediction_col = "pred_test",
                      
                      # Customizing the plot
                      add_normalized = TRUE,
                      add_col_percentages = FALSE,
                      add_row_percentages = FALSE,
                      counts_col = "n",
                      )

```

## Calculate some metrics out of the confusion matrix

**ACCURACY:** How many observations did we get right?

```{r}
accuracy <- sum(pred_test == target_test)/nrow(test_samples)
cat(sprintf("%.2f %%", accuracy*100)) # cat is just another way to print() things
```

**SPECIFICITY:** True Negative Rate

```{r}
total_real_no      <- sum(target_test == "No")
total_correct_no   <- sum(pred_test == "No" & target_test == "No")

TNR <- total_correct_no/total_real_no
cat(sprintf("%.2f %%", TNR*100))

```

**SENSITIVITY:** True Positive Rate

```{r}
total_real_yes      <- sum(target_test == "Yes")
total_correct_yes   <- sum(pred_test == "Yes" & target_test == "Yes")

TPR <- total_correct_yes/total_real_yes
cat(sprintf("%.2f %%", TPR*100)) 

```


**ðŸŽ¯ ACTION POINT 04:** Go back up and tweak the `selected_threshold` until you find one that works "better" for your **test data**!

**ðŸŽ¯ ACTION POINT 05:** Once the previous step is done, paste the following on the designated space on Slack:

``` r
Spokesperson <- "<Replace this by the name of the spokesperson for your group>"
selected_threshold <- <Replace by your selected value>

## Answer the following about your training data
accuracy <- <Replace by the measure of accuracy you obtained>
TNR <- <Replace by the measure of TNR you obtained>
TPR <- <Replace by the measure of TPR you obtained>

## Answer the following about your test data
accuracy <- <Replace by the measure of accuracy you obtained>
TNR <- <Replace by the measure of TNR you obtained>
TPR <- <Replace by the measure of TPR you obtained>
```

# Step 6 (If we have the time):

If we have time, we will see which model is best at predicting the external set!

**ðŸŽ¯ ACTION POINT 06:** Feel free to back to Step 5 above and fit a new logistic regression with different features, interaction effects etc. See if any modification improves your `model`, both in the training set as well as the test set. Remember to tweak the threshold again.


**ðŸŽ¯ ACTION POINT 07:** Use the code below to save your model to your disk, then upload the `rds` file to the designated space on Slack

```{r}
model$threshold <- selected_threshold
saveRDS(model, file="ml_model_group_<SPOKESPERSON>.rds")
```

I will run your model on the external data and report back to you the results!
