# Step 3: Support Vector Machines (SVM) (25 mins)

## Step 3.1: Train a Support Vector Machine (SVM) (10 mins)

Since we know the data we are modelling has a nonlinear relationship, let's train a **Support Vector Machine** (SVM) with a radial, or Gaussian, kernel. R does not come with SVM, so we need to import it from a library. Here we will use the function `svm` from the [`e1071` package](https://cran.r-project.org/web/packages/e1071/e1071.pdf).

Here are a few things to know about the `svm` function:

- The `svm` command is largely similar to other commands such as `lm` and `glm` in that the first parameter is a formula and the second is the data set. 
- There are a few other options, but we will focus on specifying a radial kernel using `kernel = 'radial'`. 
- We _need_ to specify the type of machine learning task we are performing. Since we are doing classification, we use the option `type = 'C-classification'`.

```{r}
library(e1071)

svm.model <- svm(is_chinstrap ~ ., 
                 data = penguins_cleaned,
                 type = 'C-classification',
                 kernel = 'radial')
```

Let's look at the summary of the svm model:

```{r, output=FALSE}
summary(svm.model)
```

🎯 **ACTION POINT**: What does the 'Number of Support Vectors' represent? 

## Step 3.2: Goodness-of-Fit of the SVM (5 mins)

Let's investigate how well our model fits the data. Let's reuse the model we trained (`svm.model`) and predict the _same samples_ we used to train it. To avoid modifying our original dataframe, let's save the output of the prediction in an auxiliary df (`plot_df`):

```{r, output=FALSE}
plot_df <- 
    penguins_cleaned %>% 
    mutate(class_pred = predict(svm.model, newdata = .),
           correct    = class_pred == is_chinstrap)
plot_df
```

Remember what we did in the Decision Tree example!


Simple confusion matrix:

```{r, output=FALSE}
confusion_matrix <- 
    table(expected=plot_df$is_chinstrap, class_pred=plot_df$class_pred)
print(confusion_matrix)
```

Nicer looking confusion matrix:

```{r, output=FALSE}
library(cvms)

plot_confusion_matrix(as_tibble(confusion_matrix), 
                      target_col = "expected", 
                      prediction_col = "class_pred",
                      
                      # Customizing the plot
                      add_normalized = TRUE,
                      add_col_percentages = FALSE,
                      add_row_percentages = FALSE,
                      counts_col = "n",
                      )
```

## Step 3.3: Visualize the SVM decision space (10 mins)

Here we will demonstrate how you could simulate some data to cover the entire feature space of the data we are modelling. What do we mean by that? By inspecting `bill_length_mm` and `bill_depth_mm`, we see the range values these variables can assume: 

```{r}
penguins_cleaned %>% 
    select(c(bill_length_mm, bill_depth_mm)) %>%
    summary()
```

💡 We can simulate data to account for all possible combinations of `bill_length_mm` and `bill_depth_mm`. We achieve this using `crossing`, another `tidyverse` function:

- We feed `crossing` a sequence of numbers that range from the minimal and maximal values of both variables, incremented by 0.1 values.
- We then create a new variable `class_pred` which uses the SVM model object to predict whether a given row will be a Chinstrap penguin or not.
- Note that we say `newdata = .` to indicate that we simply want to use the data set created with `crossing` as our new data.

```{r}
sim.data <- 
  crossing(bill_length_mm = seq(31,60,0.1),
           bill_depth_mm = seq(13,22,0.1)) %>% 
  mutate(class_pred = predict(svm.model, newdata = .))
head(sim.data)
```

The data above is all synthetic ("fake")! But it is very useful to colour the background of our plot.

We use `geom_tile` to show the area the SVM model identifies as Chinstrap penguins. We then use `geom_point` to overlay the actual data. Red dots in blue areas and vice versa indicate cases where the SVM model makes errors.

```{r, dpi=300}
library(ggsci) # Just to use fancy colours.

plot_df %>%   
  ggplot() +

  # Tile the background of the plot with SVM predictions
  geom_tile(data = sim.data, 
    aes(x=bill_length_mm, y=bill_depth_mm, fill = class_pred), 
        alpha = 0.25) +

  # Actual data
  geom_point(
    aes(x=bill_length_mm, y=bill_depth_mm, colour = is_chinstrap, shape = correct), 
    size=1.5, stroke=0.8, alpha=0.8) +

  # Define X and Os
  scale_shape_manual(values = c(4, 1)) +
  
  # (OPTIONAL) Customizing the colours and theme of the plot
  scale_colour_lancet() +
  scale_fill_lancet() +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        legend.position = 'bottom', 
        plot.title = element_text(hjust = 0.5)) +
  labs(x = 'Bill Length (mm)', y = 'Bill Depth (mm)', 
       fill = 'Is Chinstrap?', colour = 'Is Chinstrap?', 
       shape = 'Correct prediction?',
       title = sprintf('Overall Training Accuracy = %.2f %%', 
                       100*(sum(plot_df$correct)/nrow(plot_df))))
```

🤝 **WORKING TOGETHER** In pairs, discuss what you see in the plot:

- What do the shape of dots represent? The `X` and `O`s?
- What do the colours of the dots represent?
- What do the background colour in the plot represent?
- Can you point in the plot roughly which dots you would expect to be the support vectors?