---
title: "Week 03 - Lab Roadmap - Linear Regression"
author: "Yijun Wang"
date: "2022-10-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simple and Multiple Linear Regression

This week, we will fit simple and multiple linear regression models in R and learn to interpret the R output.

---


# Step 1 : First Analyse: Simple linear regression (15 Min)

### Loading the `ISLR2` Package

1. Install and load the `ISLR2` package, which contains a large collection of data sets and functions.

```{r ISLR2, message = FALSE}
# install.packages("ISLR2", repos = list(CRAN="http://cran.rstudio.com/"))
library (ISLR2)
```
<br>

### Dataset
2. Use the `Boston` data set in the `ISLR2` library. It records `medv` (median house value) for 506 census tracts in Boston. Have a look at the first few rows of the Boston data set:

```{r Boston}
head (Boston)
```

We want to predict `medv` using the available predictors, such as `rm` (average number of rooms per house), `age` (average age of houses), and `lstat` (percentage of households with low socioeconomic status).  

<br>


### Linear Regression

3. Fit a simple linear regression `lm()` model, with `medv` as the response and `lstat` as the predictor:

```{r simple_lm}
lm.fit <- lm(medv ~ lstat , data = Boston)

```

The basic syntax is `lm(y ∼ x, data)`, where `y` is the response, `x` is the predictor, and `data` is the data set in which we keep these two variables.

<br>


4. Use the `tidy` function to create a dataframe with columns for the estimate, standard error, f-statistic (estimate/standard error), p-values, and 95 percent confidence intervals:

``` {r simple_summary}
library(broom)

tidy(lm.fit, conf.int = TRUE)

```

Because `lm.fit` is a simple linear regression model, there are only two coefficients: $\hat{\beta}_0$ and $\hat{\beta}_1$. The goodness-of-fit of the model can be measured by the $R^2$ in the output, which can be obtained (along with other model statistics) using the `glance` function.

```{r simple_coefficients}
glance(lm.fit)$r.squared
```
<br>

### Plotting Results
7. Plot `medv` and `lstat` along with the least squares regression line using the `geom_point()` and `geom_abline()` functions.:

```{r simple_plotting,message=FALSE}
library(tidyverse)

ggplot(data = Boston, aes(x = lstat, y = medv)) +
  geom_point() +
  geom_abline(intercept = lm.fit$coefficients[1], slope = lm.fit$coefficients[2])
```
<br>


---

# B) Step 2: Second Analyse:  Multiple linear regression
<br>


We will still use the `Boston` data set to fit multiple linear regression. The fitting process is similar to simple linear regression.
<br>


### Linear Regression
1. Fit a multiple linear regression `lm()` model, with `medv` as the response, `lstat` and `age` as the predictors:

```{r multi_lm}
lm.fit <- lm(medv ~ lstat + age , data = Boston)
```

```{r multi_Summary}
tidy(lm.fit, conf.int = TRUE)
```

The syntax `lm(y ~ x1 + x2 + x3)` is used to fit a model with three predictors, `x1`, `x2`, and `x3`. The `tidy()` function now outputs the regression coefficients for all the predictors.
<br>


2. Fit a multiple linear regression `lm()` model, with `medv` as the response, all rest variables as the predictors:

```{r multi_all_lm}
lm.fit <- lm(medv ~ ., data = Boston)
tidy(lm.fit, conf.int = TRUE)
```
We can access the individual components of a summary object by name (type `?glance` to see what is available). Hence `glance(lm.fit)$r.squared` gives us the $R^2$.
<br>


### Variable Selection
<br>

3. Select variables:

In these two multiple linear regression models, the t-tests and F-test results suggest that many of the predictors are significant for the response variable. However, some do not achieve statistical significance. Can you see which variables these are? 

We call the process of determining which predictors are associated with the response as **variable selection**.

If the number of predictors is very small, we could perform the variable selection by trying out a lot of different models, each containing a different subset of the predictors. We can then select the best model out of all of the models we have considered.

Using the template below, try figuring out the model which produces the highest adjusted $R^2$. The adjusted $R^2$ has a similar interpretation to $R^2$, only it is an advantage here as it penalises models that include insignificant parameters.


```{r variable_selection}

lm.fit <- lm(medv ~ ., data = Boston)

glance(lm.fit)$adj.r.squared

```
We found that if you remove `indus` and `age`, the adjusted $R^2$ becomes slightly larger compared to including all predictors.

```{r variable_selection_after}

lm.fit <- lm(medv ~ ., data = Boston[,-c(3,7)])

glance(lm.fit)$adj.r.squared

```

---


# Step 3:  Some potential problems (15 min)
<br>



Many problems may occur when we fit a linear regression model to a particular data set. These problems will lead to inaccurate estimation. In this step, we will identify and overcome potential problems such as outliers, collinearity and interaction effects. 
<br>


### 1. Handle interaction terms:

In regression, an interaction effect exists when the effect of an independent variable on the response variable changes, depending on the values of one or more independent variables. When you believe there is an interaction effect, it is easy to include interaction terms in a linear model using the `lm()` function.

```{r interaction}
tidy(lm(medv ~ lstat * age , data = Boston))
```

 The syntax `lstat:age` tells R to include an interaction term between lstat and age The syntax `lstat*age` simultaneously includes `lstat`, `age`, and the interaction term `lstat×age` as predictors; it is a shorthand for `lstat+age+lstat:age`.
 <br>


### 2. Identify outliers through residual plots:
An outlier is a point for which $\hat{y}_i$ is far from the value predicted by the model. Outliers can arise for a variety of reasons, such as incorrect recording of observation during data collection. Outliers could be identified through residual plots:

```{r outlier}
par(mfrow = c(2, 2))
plot(lm.fit)
```
The `plot` function automatically produces four diagnostic plots when you pass the output from `lm()`. Plots on the left column are residual plots, indicating the relationship between residuals and fitted values.

In practice, it can be difficult to decide how large a residual needs to be before we consider the point to be an outlier. Instead of plotting the residuals, we can address this problem by plotting the studentized residuals. These are computed by dividing each residual e<sub>i</sub> by its estimated standard studentized residual error. Observations with studentized residuals greater than 3 in absolute value are possible outliers. Using the plot() function to plot the studentized residuals:

```{r outlier_plot}
plot(predict(lm.fit), rstudent(lm.fit))

```
<br>




### 3. Handle outliers:
If we believe an outlier is due to an error in data collection or recording, then one solution is to simply remove the observation. However, care should be taken, as an outlier may instead signal a deficiency with our model, such as a missing predictor.
<br>



### 4. Detect multicollinearity using the correlation matrix:
Multicollinearity refers to the situation in which two or more predictor variables are highly correlated to one another. It can be detected through the correlation matrix:
```{r cor_boston}
cor(Boston)
```
Ignoring the last row and the last column in the matrix, which indicate the relationship with response variable `medv`, an element of this matrix that is large in absolute value indicates a pair of highly correlated variables, and therefore a collinearity problem in the data.

We can detect multicollinearity quantitatively using `vif()` function in the `car` package:

```{r installing car and calling vif, message = FALSE }

#install.packages("car", repos = list(CRAN="http://cran.rstudio.com/"))
library(car)

```


```{r cor_lm , message = FALSE}
vif(lm.fit)

```

Instead of inspecting the correlation matrix, a better way to assess multicollinearity is to compute the variance inflation factor (VIF). As a rule of thumb, a VIF value that exceeds 5 or 10 indicates a problematic amount of collinearity.
<br>



### 5. Handle collinearity:

When faced with the problem of multicollinearity, there are two simple solutions. 

-The first is to drop one of the problematic variables from the regression.

-The second solution is to combine the collinear variables into a single predictor, where such combination makes theoretical sense.
<br>

---



# Step4: Practical exercises (50 min)

So far, we have learnt to fit simple and multiple linear regression models in R. In this practical case, we will continue to use the data set `Auto` studied in the last lab. Make sure that the missing values have been removed from the data.

Eight questions are listed below. You are required to try to answer these questions in pairs using R commands. We will go over the solutions once everyone has finished these questions.
<br>


### Questions

1. Use the `lm()` function to perform a simple linear regression with `mpg` as the response and `horsepower` as the predictor. Use the `tidy()` function to print the results. Comment on the output. For example:

   i. Is there a relationship between the predictor and the response?
   ii. How strong is the relationship between the predictor and the response?
   iii. Is the relationship between the predictor and the response positive or negative?
   iv. What is the predicted `mpg` associated with a `horsepower` of 98? What are the associated 95 % confidence intervals?

   
2. Plot the response and the predictor. Use the `geom_abline()` function to display the least squares regression line.

3. Use the `plot()` function to produce diagnostic plots of the least squares regression fit. Comment on any problems you see with the fit.

4. Produce a scatterplot matrix that includes all the variables in the data set.

5. Compute the matrix of correlations between the variables using the function `cor()`. You will need to exclude the `name` and `origin` variable, which is qualitative.

6. Use the `lm()` function to perform a multiple linear regression with `mpg` as the response and all other variables except `name` as the predictors. Use the `tidy()` function to print the results. Comment on the output. For instance:
   i. Is there a relationship between the predictors and the response?
   ii. Which predictors appear to have a statistically significant relationship to the response?
   iii. What does the coefficient for the year variable suggest?
   

7. Use the `plot()` function to produce diagnostic plots of the linear regression fit. Comment on any problems you see with the fit. Do the residual plots suggest any unusually large outliers? 

8. Use the `*` and `:` symbols to fit linear regression models with interaction effects. Do any interactions appear to be statistically significant?



